---
title: "Chapter 2: Bayesian Methods and WinBUGS"
author: "Bob"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')

setwd("~/Documents/projects/Healthcare_Technology_Assessment/ARSCESDMH")
```

# `BLOCKER.txt`

## BLOCKER EG - Modelling log(Odds Ratio) 

```{r blocker_initial_model}


# Model

model_code <- "model {
  # Likelihood
  for (j in 1:Nstud) {
    P[j] <- 1/V[j]      # Calculate precision
    Y[j] ~ dnorm(delta[j],P[j])
    delta[j] ~ dnorm(d,prec)
  }
  
  # Priors
  d ~ dnorm(0,1.0E-6)
  prec <- 1/tau.sq
  tau.sq <- tau*tau   # tau.sq = between-study variance
  tau ~ dunif(0,10)   # Uniform on SD
}" %>% textConnection


# Starting/initial values

initial_values <- list(list(
  delta=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0),
  d=0,
  tau=1
))


# Data
# The data is inline below, but if you want to read it from the original excel file, you could do this:
# library('readxl')
# data <- read_excel("BLOCKER.xls") %>% as.list
# data['Nstud'] <- 22

data <- "Y	V
0.02817088	0.72301587
-0.7410032	0.23343543
-0.5406212	0.31872921
-0.24612808	0.01909462
0.06945337	0.07876801
-0.5841569	0.45658762
-0.51238549	0.01923431
-0.07862326	0.04161232
-0.42417337	0.07506123
-0.3348234	0.01370499
-0.21339753	0.0379751
-0.03890844	0.05265218
-0.59325371	0.18076732
0.28154593	0.04220784
-0.32133359	0.08863069
-0.13534792	0.06808023
0.14060645	0.13262283
0.32204972	0.30541643
0.44438052	0.51358597
-0.21750973	0.0675177
-0.59107599	0.06615537
-0.60809913	0.07419013" %>% 
  read.delim(text=., sep='\t') %>%
  as.list

data['Nstud'] <- length(data[['V']])

out_initial <- jags(data = data,
            inits = initial_values,
            parameters.to.save = c("d", "tau"),
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 30000,
            n.burnin = 20000,
            n.thin = 2)

plot(out_initial)

```

# `BLOCKER_binomial.txt`

BLOCKER EG - Modelling binomial data, as per WinBUGS examples, vol I
Note: some of the variance/precision parameter names are different to the code in the WinBUGS example blocker file.


```{r BLOCKER_binomial}
model_code <- "
model {
  for( j in 1 : Nstud ) {
    # Binomial Likelihood
    rc[j] ~ dbin(pc[j], nc[j])
    rt[j] ~ dbin(pt[j], nt[j])
    logit(pc[j]) <- mu[j]
    logit(pt[j]) <- mu[j] + delta[j]   # delta[j] is the log odds ratio in jth study
    mu[j] ~ dnorm(0.0,1.0E-5)
    delta[j] ~ dnorm(d, prec)      # d = pooled log odds ratio & prec = between-study precision
  }

  # Priors
  d ~ dnorm(0.0,1.0E-6)
  prec <- 1/tau.sq
  tau.sq <- tau*tau        # tau = between-study standard deviation
  tau ~ dunif(0,10)          # Uniform prior on SD
  
  # Calculate overall pooled Odds Ratio (OR)
  OR <- exp(d)
  
  # Calculate probability that the pooled OR > 1
  prob.OR1 <- step(d)
}
" %>% textConnection

```

```{r how_to_get_started}

# Data
data <- list(
  rt = c(3,  7,  5,  102,  28, 4,  98,  60, 25, 138, 64, 45,  9, 57, 25, 33, 28, 8, 6, 32, 27, 22 ),
  nt = c(38, 114, 69, 1533, 355, 59, 945, 632, 278,1916, 873, 263, 291, 858, 154, 207, 251, 151, 174, 209, 391, 680),
  rc = c(3, 14, 11, 127, 27, 6, 152, 48, 37, 188, 52, 47, 16, 45, 31, 38, 12, 6, 3, 40, 43, 39),
  nc = c(39, 116, 93, 1520, 365, 52, 939, 471, 282, 1921, 583, 266, 293, 883, 147, 213, 122, 154, 134, 218, 364, 674),
  Nstud = 22)

initial_values <- list(
  list(
    d = 0.5,
    tau=0.5,
    mu = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
    delta = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0)
  )
)

for (i in 1:3){
  new_inits <- list(
    d = 0,
    tau=1,
    mu = runif(data$Nstud),
    delta = runif(data$Nstud)
  )
  
  initial_values <- append(initial_values, list(new_inits))
}

results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = c("OR", "d", "tau"),
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 60000,
            n.burnin = 30000,
            n.thin = 2)

summary(results)
plot(results)
```
# `blocker_with_results.txt`

## BLOCKER EG - Modelling log(Odds Ratio)

Using multiple chains to assess convergence

```{r blocker_eg}

model_code <- "
model {
  # Likelihood
  for (j in 1:Nstud) {
    P[j] <- 1/V[j]
    Y[j] ~ dnorm(delta[j],P[j])
    delta[j] ~ dnorm(d, prec)
  }
  
  # Priors
  d ~ dnorm(0,1.0E-6)
  prec <- 1/tau.sq
  tau.sq <- tau*tau
  tau ~ dunif(0,10)
  
  # Calculate OR & prob OR>1
  OR <- exp(d)
  prob.OR1 <- step(d)
  
  #Set delta[26] = d for caterpillar plot
  delta[26] <- d
}" %>% textConnection

initial_values <- list(
  # Chain 1
  list(delta=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0, NA,NA,NA,NA),
       d=0,
       tau=1),

  # Chain 2
  list(delta=c(10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10,10, NA,NA,NA,NA),
       d=-10,
       tau=8)
)

data <- list(
  Y=c(0.02817088,-0.74100320,-0.54062120,-0.24612808,0.06945337,-0.58415690,
    -0.51238549,-0.07862326,-0.42417337,-0.33482340,-0.21339753,-0.03890844,
    -0.59325371,0.28154593,-0.32133359,-0.13534792,0.14060645,0.32204972,
    0.44438052,-0.21750973,-0.59107599,-0.60809913),
  V=c(0.72301587,0.23343543,0.31872921,0.01909462, 0.07876801, 0.45658762,
      0.01923431, 0.04161232, 0.07506123, 0.01370499, 0.03797510, 0.05265218,
      0.18076732, 0.04220784, 0.08863069, 0.06808023, 0.13262283, 0.30541643,
      0.51358597, 0.06751770, 0.06615537, 0.07419013),
  Nstud=22
)

out_eg <- jags(data = data,
            inits = initial_values,
            parameters.to.save = c("OR", "d", "tau", "delta"),
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 30000,
            n.burnin = 20000,
            n.thin = 2)

plot(out_eg)

# densityplot(out)
```

These are the comments from the results of the WinBUGS file:

"The summary statistics for the 2 chains using a burn-in of 20,000 iterations 
and a sample of length 10,000 iterations are very similar despite the very 
different starting values suggesting that the chains have converged.

The BGR statistics are also suggestive of the chains having converged - the 
green (pooled) and blue (individual chains) widths of the 80% CrIs are relatively 
stable, and their ratio (red line) is approximately 1.

The actual samples from the 2 chains are also very similar as can be 
seen on the history plots below for d and tau."

## TO DO: 

How do we make a caterpillar plot from the delta and d values? Note that the model copied the value for 'd' into the delta vector (`delta[26] <- d`), which was a hack to put all of these values in the same vector for plotting. We might not need to do this.

We want these values ordered by name; they seem to be ordered by decreasing value of the median.

```{r caterplot_delta}

colnames <- sprintf("delta[%d]", c(1:22, 26))
caterplot(out_eg$samples[, colnames])


colnames <- c(sprintf("delta[%d]", 1:22), 'd')
caterplot(out_eg$samples[, colnames]) 
```



