---
title: 'Chapter 8: Multi-parameter evidence synthesis'
author: "Bob"
date: "2025-02-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')
```

# `maple_syrup_example`

CONFIDENCE PROFILE EXAMPLE: Maple Syrup Urine Disease

__Objective__: To estimate the probability of retardation for a case of MSUD without screening (theta.n), & change in retardation rate associated with screening (e.d)


```{r maple_syrup_example}


# Model

model_code <- "model
{

  #All data assumed to arise from binomial distributions with the appropriate parameters
   
   r.r ~ dbin(r, n.r)
   r.s ~ dbin(phi.s, n.s)
   r.n ~ dbin(phi.n, n.n)
   r.em ~ dbin(theta.em, n.em)
   r.lm ~ dbin(theta.lm, n.lm)  

  #Define functional relationships
	# Prob retardation for a case of MSUD who is screened
  	 theta.sm <- phi.s * theta.em + (1 - phi.s)*theta.lm

	# Prob retardation for a case of MSUD who is NOT screened				
  	 theta.nm <- phi.n * theta.em + (1 - phi.n) * theta.lm	

 	# Expected retardation per 100000 newborns who are screened
  	 theta.s <- (theta.sm * r) * 100000

 	# Expected retardation per 100000 newborns who are NOT screened							
  	 theta.n <- (theta.nm * r) * 100000

   
  # Change in retardation rate associated with screening

  e.d <- theta.s - theta.n


  # Prior distributions - 'non-informative' Beta(1,1) priors

   r ~ dbeta( 1, 1)
   phi.s ~ dbeta( 1, 1)
   phi.n ~ dbeta( 1, 1)
   theta.em ~ dbeta( 1, 1)
   theta.lm ~ dbeta( 1, 1)

}
" %>% textConnection


# Data

data <- list( 
  n.r = 724262, 
  r.r = 7, 
  n.s = 276, 
  r.s = 253, 
  n.n = 18, 
  r.n = 8,  
  n.em=10, 
  r.em = 2, 
  n.lm = 10, 
  r.lm = 10)


# Starting/initial values



parameters_to_save <- c("e.d") # What else?


results <- jags(data = data,
            # inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = 1, # length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

plot(results)
```

## `hivepi_6`

Simple version of the HIV Epi model with just 6 data points.


```{r hivepi_6}


# Model

model_code <- "model{
#  SET PRIORS   
   a ~ dbeta( 1,2)            
   z ~ dbeta (1,1)      
   b <- z * (1-a)             #  sets constraint (1-a-b > 0)       
   c ~ dbeta (1,1)             
   d ~ dbeta (1,1)             
   e ~ dbeta (1,1)            

# VECTOR p[1:6] HOLDS THE EXPECTED PROBABILITIES FOR EACH DATA POINT
   p[1] <- a
   p[2] <- b
   p[3] <- c
   p[4] <- d
   p[5] <- (d*b + e*(1-a-b))/(1- a)
   p[6] <- c*a + d*b + e*(1-a-b) 
    
#  LIKELIHOOD AND DIAGNOSTICS
   for(i in 1: 6) {
       r[i] ~ dbin(p[i],n[i])                                                                                                      
       rhat[i] <- p[i] * n[i]                                                                                                                    
       dev[i] <- 2 * (r[i] * log(r[i]/rhat[i])  +  (n[i]-r[i]) * log((n[i]-r[i])/(n[i]-rhat[i])))    
    } 
    resdev <- sum(dev[])                                                                                                         
}
" %>% textConnection


# Data

data <- list( 
  r=c(11044, 12, 252, 10, 74, 254),
  n=c(104577, 882, 15428, 473, 136139, 102287)
)


# Starting/initial values: None specified

parameters_to_save <- c("a", "b", "c", "d", "e", "p", "r", "rhat", "dev", "resdev")


results <- jags(data = data,
            # inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = 1, # How many chains?
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

plot(results)
```

## `hivepi_6_xval`

Simple version of the HIV Epi model with just 6 data points. (Yes, it has exactly the same comment as last example. But how is this one different?)


```{r hivepi_6_xval}


# Model

model_code <- "model
{
  #  SET PRIORS   
   a ~ dbeta( 1,2)            
   z ~ dbeta (1,1)      
   b <- z * (1-a)             #  sets constraint (1-a-b > 0)       
   c ~ dbeta (1,1)             
   d ~ dbeta (1,1)             
   e ~ dbeta (1,1)            

# VECTOR p[1:6] HOLDS THE EXPECTED PROBABILITIES FOR EACH DATA POINT
   p[1] <- a
   p[2] <- b
   p[3] <- c
   p[4] <- d
   p[5] <- (d*b + e*(1-a-b))/(1- a)
   p[6] <- c*a + d*b + e*(1-a-b) 
                                              


#  LIKELIHOOD AND DIAGNOSTICS
  for(i in 1:3) {       r[i] ~ dbin(p[i],n[i])   
       rhat[i] <- p[i] * n[i]               
       dev[i] <- 2 * (r[i] * log(r[i]/rhat[i])  +  (n[i]-r[i]) * log((n[i]-r[i])/(n[i]-rhat[i])))    
}                                                                                                   

 for(i in 5:6) {       r[i] ~ dbin(p[i],n[i])   
       rhat[i] <- p[i] * n[i]               
       dev[i] <- 2 * (r[i] * log(r[i]/rhat[i])  +  (n[i]-r[i]) * log((n[i]-r[i])/(n[i]-rhat[i])))    
}        
	dev[4]<-0
    resdev <- sum(dev[])        

#cross-validation of data point 4
    r.rep ~ dbin(p[4],n[4])
    p.xval <- step(r.rep - r[4])  -   0.5 * equals(r.rep,r[4])    
                                                                              
}
" %>% textConnection


# Data (same as before)

data <- list( 
  r=c(11044, 12, 252, 10, 74, 254),
  n=c(104577, 882, 15428, 473, 136139, 102287)
)


# Starting/initial values: None specified

parameters_to_save <- c("a", "b", "c", "d", "e", "p", "r", "rhat", "dev", "resdev", "p.xval")


results <- jags(data = data,
            # inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = 1, # How many chains?
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

plot(results)
```
