---
title: 'Chapter 3: Introduction to Decision Models'
author: "Bob"
date: "2025-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_code, warn=FALSE, message=FALSE}

# libraries
library('rjags')      # coda.samples
library('jagsUI')     # wrapper for rjags
library('mcmcplots')  # caterplot
library('coda')
library('knitr')
library('dplyr')
library('tidyr')
library('ggplot2')

# functions
coda_sample_2_df <- function(my_coda_sample){
  # Extract data from a coda sample into a dataframe
  seq_along(my_coda_sample) %>% lapply(function(chain){
    df <- my_coda_sample[[chain]] %>% as.matrix %>% as.data.frame
    df['chain'] <- chain
    df
  }) %>% 
    bind_rows %>% 
    mutate(chain=factor(chain))
}

plot_densities_from_coda_df <- function(vars, coda_df){
  # Make comparative density plots for the given variables
  coda_df %>% 
    select(all_of(vars)) %>%
    pivot_longer(cols=vars) %>% 
    ggplot(aes(x=value, col=name, fill=name)) + geom_density(alpha=0.6)
}

```

# Deterministic decision tree

This is the model at the end of section 3.4 (p 56).

## Effectiveness data

From [Bibi et al., 1994](https://pubmed.ncbi.nlm.nih.gov/8051377/)

```{r effectiveness_data}

effectiveness_df <- data.frame(
  Description = c("Prophylactic antibiotics", "Placebo"),
  Infection = c(4, 28),
  No_infection = c(129, 108),
  row.names = c("Treatment", "Control")
) %>% mutate(Total = Infection + No_infection)

effectiveness_df %>% kable
```

## Resource use and cost data

From Table 3.3, p55. Data is from [Mugford et al., 1989](https://pubmed.ncbi.nlm.nih.gov/2511938/), except for the cost of administering antibiotic which was estimated by Welton et al.


```{r resource_use_and_cost_data}

cost_df <- data.frame(
  Parameter = c( "Length of stay, wound infection",
                 "Length of stay, no wound infection",
                 "Cost of inpatient stay per day, wound infection",
                 "Cost of inpatient stay per day, no wound infection",
                 "Cost of prophylactic cephaslosporin, per dose",
                 "Cost of administering the antibiotic, per dose",
                 "Doses administered",
                 "Number of Caesarian sections total",
                 "Number developing wound infections without antibiotics"
              ),
  Estimate = c(8.80, 6.70, 163.03, 107.26, 5.67, 7.00, 3, 486, 41),
  Units=c("days", "days", "£", "£", "£", "£", "count", "count", "count"),
  Variable_name=c("loswd", "losnwd", "cstwd", "cstnwd", "cstPx", "cstadmin", "dose", "nc1", "rc1")
)

cost_df %>% kable
```



```{r deterministic_decision_tree}

model_code <- "
model {
  cost_nwdpx <- losnwd * cstnwd + dose * (cstPx + cstadmin) # Cost (No infection/Px)
  cost_wdpx <- loswd * cstwd + dose * (cstPx + cstadmin)    # Cost (Infection/Px)
  cost_nwd <- losnwd * cstnwd                               # Cost (No infection/no Px)
  cost_wd <- loswd * cstwd                                  # Cost (Infection/no Px)
  
  RR <- (a/(a+b))/(c/(c+d))                                 # Relative risk using 
                                                            # data from table 3.2
                                                            
  p1 <- rc1/nc1
  p2 <- RR * p1
  
  costtrt <- ((1-p2) * cost_nwdpx) + p2 * cost_wdpx         # Total cost (payoff) Px
  costctl <- ((1-p1) * cost_nwd) + p1 * cost_wd             # Total cost (payoff) No Px

}"

cost_data <- with(cost_df, setNames(as.list(Estimate), nm=Variable_name))
effectiveness_data <-  list(
  a=effectiveness_df['Treatment', 'Infection'],
  b=effectiveness_df['Treatment', 'No_infection'],
  c=effectiveness_df['Control', 'Infection'],
  d=effectiveness_df['Control', 'No_infection']
)
deterministic_data <- append(cost_data, effectiveness_data)

# deterministic_data <- list(
#   losnwd=6.7,
#   loswd=8.8,
#   cstnwd=107.26,
#   cstwd=163.03,
#   cstPx=5.67,  # !!! `cstdrug` should be `cstPx`
#   cstadmin=7,
#   dose=3,
#   rc1=41,
#   nc1=486, 
#   a=4,
#   b=129,
#   c=28,
#   d=108
# )

results <- jags(
  data = deterministic_data,
  parameters.to.save = c("p1", "p2", "costtrt", "costctl"),
  model.file = model_code %>% textConnection,
  n.chains = 1,
  n.iter=1
)


results
```

 The expected costs of the treament and control branches match those in the text on p56.

# Stochastic decision tree

This is the model on pp 59-60.


```{r stochastic_decision_tree}

stochastic_model_code <- "
model{
  lnRR ~ dnorm(theta, prec)                  # Distribution for ln(Relative Risk)
  theta <- log( (a/(a+b)) / (c/(c+d)) )
  prec <- 1/( (1/a) - (1/(a+b)) + (1/c) - (1/(c+d)) )
  
  p1 ~ dbeta(alpha, beta)                    # Distribution for Prob(Infection/NoPx)
  alpha <- rc1
  beta <- nc1 - rc1
  
  p2 <- exp(lnRR) * p1                       # Distribution for Prob(Infection/Px)
  
  loswd ~ dnorm(mnloswd, precwd)             # Distribution for length of stay with infection
  precwd <- 1/pow(sdloswd/sqrt(numwd), 2)
  
  losnwd ~ dnorm(mnlosnwd, precnwd)          # Distribution for length of stay w/o infection
  precnwd <- 1/pow(sdlosnwd/sqrt(numnwd), 2)
  
  cstadmin ~ dunif(4, 10)                    # Px administration
  
  cst.trt <- (1-p2)*((cstPx + cstadmin)*3 + (losnwd*cstnwd)) + p2*((cstPx + cstadmin)*3 + (loswd*cstwd)) # Total cost (payoff) Px
  
  cst.ctl <- (1-p1)*(losnwd*cstnwd) + p1*(loswd*cstwd) # Total cost (payoff) No Rx
  
  diff.cost <- cst.trt - cst.ctl             # Difference in cost
}" %>% textConnection()

stochastic_data <- list(
  rc1=41, nc1=486, cstwd=163.03, cstnwd=107.26,
  mnloswd=8.8, sdloswd=3.5,
  mnlosnwd=6.7, sdlosnwd=7.1,
  numwd=41, numnwd=445, # !!! numwd == rc1; numnwd == (nc1 - rc1)
  cstPx=5.67,
  # rt=4, nt=133, rc=28, nc=136,  # !!! Unused variables
  a=4, b=129, c=28, d=108
)

# Here I code the changes from the data for the deterministic model to the 
# data for the stochastic one, in case that makes the differences easier to see.

    # Add these:
    # setdiff(names(stochastic_data), names(deterministic_data))
    # "mnloswd"  "sdloswd"  "mnlosnwd" "sdlosnwd"
    # "numwd" ==  rc1;
    # "numnwd" == (nc1 - rc1)
    # 
    # Remove or rename these:
    # setdiff(names(deterministic_data), names(stochastic_data))
    #  "loswd" -> "mnloswd"
    #  "losnwd" -> "mnlosnwd"
    #  "cstadmin" hard-coded uniform distribution
    #  "dose": hard-coded !!!

# uncertainty <- c(sdloswd=3.5, sdlosnwd=7.1) # standard deviations from Table 3.4
# rename_me <- c("loswd"="mnloswd", "losnwd"="mnlosnwd")
# 
# stochastic_data <- append(deterministic_data, uncertainty)
# for (old_name in names(rename_me)){
#   new_name <- rename_me[[old_name]]
#   names(stochastic_data)[names(stochastic_data) == old_name] <- new_name
# }
# stochastic_data["cstadmin"] <- NULL # now coded as uniform random
# stochastic_data["dose"] <- NULL # the 3 is hard coded now
# 
# stochastic_data["numwd"] <- stochastic_data["rc1"]; 
# stochastic_data["numnwd"] <-  with(stochastic_data, (nc1 - rc1))


parameters_to_save <- c("cst.trt", "cst.ctl", "diff.cost")

stochastic_results <- jags( data = stochastic_data,
                            parameters.to.save = parameters_to_save,
                            model.file = stochastic_model_code,
                            n.chains = 4,
                            n.adapt = 100,
                            n.iter = 50000,
                            n.burnin = 20000)

stochastic_results_summary <- summary(stochastic_results) %>% as.data.frame()

stochastic_results_summary

```


```{r plot_stochastic_results}
plot(stochastic_results)
```


### Prior distributions


```{r prior_distributions}

with(stochastic_data, {
  curve(dbeta(x, rc1, nc1 - rc1), 
        from = 0, to = 1,
        n=1001,
        xlab = "p1", 
        ylab = "Density", 
        main = "Prior distribution for p1");
  
  curve(dnorm(x, mean=mnloswd, sd=sdloswd/sqrt(numwd)),
        from=0, to=21,
        n=1001,
        xlab = "Length of stay",
        ylab = "Density",
        main = "Prior distribution for length of stay, wound infection");
  
  curve(dnorm(x, mean=mnlosnwd, sd=sdlosnwd/sqrt(numwd)),
        from=0, to=21,
        n=1001,
        xlab = "Length of stay",
        ylab = "Density",
        main = "Prior distribution for length of stay, no wound infection");
})

```

### Simulation results

Here we re-sample an exissting model and plot densities of several variables using the R functions defined at the beginning of this document.

```{r plot_simulation_results}
my_vars <- c('p1', 'p2', 'cst.trt', 'cst.ctl', 'diff.cost')
coda_df <- coda.samples(stochastic_results$model, variable.names = my_vars, 
    n.iter = 10000) %>% 
  coda_sample_2_df %>% 
  sample_frac(1L) # randomly scramble chain order

plot_densities_from_coda_df(c('cst.trt', 'cst.ctl'), coda_df)

coda_df %>% ggplot(aes(x=diff.cost)) + geom_density()

plot_densities_from_coda_df(c('p1', 'p2'), coda_df)

```


## Economic evaluation

Here we add the code from p65 to the model above.

```{r economic_evaluation}
economic_model_code <- "
model{
  lnRR ~ dnorm(theta, prec)                  # Distribution for ln(Relative Risk)
  theta <- log( (a/(a+b)) / (c/(c+d)) )
  prec <- 1/( (1/a) - (1/(a+b)) + (1/c) - (1/(c+d)) )
  
  p1 ~ dbeta(alpha, beta)                    # Distribution for Prob(Infection/NoPx)
  alpha <- rc1
  beta <- nc1 - rc1
  
  p2 <- exp(lnRR) * p1                       # Distribution for Prob(Infection/Px)
  
  loswd ~ dnorm(mnloswd, precwd)             # Distribution for length of stay with infection
  precwd <- 1/pow(sdloswd/sqrt(numwd), 2)
  
  losnwd ~ dnorm(mnlosnwd, precnwd)          # Distribution for length of stay w/o infection
  precnwd <- 1/pow(sdlosnwd/sqrt(numnwd), 2)
  
  cstadmin ~ dunif(4, 10)                    # Px administration
  
  cst.trt <- (1-p2)*((cstPx + cstadmin)*3 + (losnwd*cstnwd)) + p2*((cstPx + cstadmin)*3 + (loswd*cstwd)) # Total cost (payoff) Px
  
  cst.ctl <- (1-p1)*(losnwd*cstnwd) + p1*(loswd*cstwd) # Total cost (payoff) No Rx
  
  diff.cost <- cst.trt - cst.ctl             # Difference in cost

  
  # Economic evaluation code from pp 65-66
  
  totQALYs.wd <- ((QALYwd/365)*loswd) + ((Fullhealth/365)*(fllwupdays-loswd))     # QALYs (infection)
  totQALYs.nwd <- ((QALYnwd/365)*losnwd) + ((Fullhealth/365)*(fllwupdays-losnwd)) # QALYs (No infection)

  QALYs.trt <- (1-p2) * totQALYs.nwd + p2 * totQALYs.wd  # QALYs (Px)
  QALYs.ctl <- (1-p1) * totQALYs.nwd + p1 * totQALYs.wd  # QALYs (no px)
  diff.QALYs <- (QALYs.trt - QALYs.ctl)                  # Difference in QALYs
  
  for (k in 1:M)
  {
  	lambda[k] <- (k-1) * 2000
  	INB[k] <- lambda[k] * diff.QALYs - diff.cost  # !!! not `delta.Qalys` or `delta.cost`
  	ProbCE[k] <- step(INB[k])
  }
}
" %>% textConnection()

# economic_data <- list(
#   rc1=41, nc1=486, cstwd=163.03, cstnwd=107.26,
#   mnloswd=8.8, sdloswd=3.5,
#   mnlosnwd=6.7, sdlosnwd=7.1,
#   numwd=41, numnwd=445,
#   cstPx=5.67,
#   a=4, b=129, c=28, d=108,
#   # additional data for economic evaluation
#   M=21, QALYwd=0.68, QALYnwd=0.88, Fullhealth=1, fllwupdays=20
# )

economic_data <- append(stochastic_data,
  # additional data for economic evaluation
  list(M=21, QALYwd=0.68, QALYnwd=0.88, Fullhealth=1, fllwupdays=20)
)

parameters_to_save <- c(#"ProbCE", 
  "cst.trt", "cst.ctl", "diff.QALYs", "diff.cost", "p1", "p2")


economic_results <- jags(data = economic_data,
            parameters.to.save = parameters_to_save,
            model.file = economic_model_code,
            n.chains = 1,
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

economic_results_summary <- summary(economic_results) %>% as.data.frame()

economic_results_summary

```
## Stochastic economic decision model results

### Cost-Effectiveness
Figure 3.4: Cost-effectiveness plane (10000 simulations) for Caesarean section example.
 
```{r plot_ce}
ce_vars <- c("diff.QALYs", "diff.cost")
ce_df <- coda.samples(economic_results$model, variable.names = ce_vars, 
    n.iter = 10000) %>% coda_sample_2_df
  
ce_df %>%
  ggplot(aes(x=diff.QALYs, y=diff.cost, col=chain)) + 
  geom_point(size=0.5, alpha=0.1) +
  geom_abline(slope=0, intercept=0, linetype='solid') +
  geom_abline(slope=10000*(1:3), intercept=0, linetype='dotted') +
  theme(legend.position="none") +
  labs(x = "Incremental QALYs", y = "Incremental costs", title="Cost-effectiveness plane for Caesarian section example")
```


### Cost-effectiveness

```{r economic_results}
inb_vars <- c("INB", "ProbCE")
inb_df <- coda.samples(economic_results$model, variable.names = inb_vars, 
    n.iter = 10000) %>% 
  coda_sample_2_df %>%
  group_by(chain) %>%
  mutate(iteration=row_number()) %>%
  ungroup() %>%
  as.data.frame

inb_df_long <- inb_df %>% 
  pivot_longer(
    cols=1:42,
    names_to=c("metric", "k"),
    names_pattern="(INB|ProbCE)\\[(\\d+)\\]",
    values_to="value"
  ) %>%
  mutate(k=as.integer(k), lambda=2000*(k-1))
```

### Cost-effectiveness acceptability curve for Caesarian section example.

Figure 3.5 [also 3.6(c)]. WTP = "Willingness To Pay"

```{r CE_acceptability_curve, message=FALSE}

inb_df_long %>%
  group_by(metric, lambda) %>%
  filter(metric=="ProbCE") %>%
  summarize(P_CE=mean(value)) %>% 
  ggplot(aes(x=lambda, y=P_CE, group=metric)) + geom_line() + 
    ylim(0,1) + 
    ylab("Probability cost effective)") + 
    xlab("Lambda, WTP per additional QALY") +
    ggtitle("Cost-effectiveness acceptability curve")
  

```

### Expected net benefit (INB) and 95% interval for a range of lambda values.

Figure 3.6(b): Expected incremental net benefit and 95% nterval for a range of lambda.

```{r expected_INB}

 # quantile(probs = c(2.5, 50, 97.5)/100)
inb_df_long %>%
  group_by(metric, lambda) %>%
  filter(metric=="INB") %>%
  summarize(
    mean_INB=mean(value),
    lo_end=quantile(value, probs=0.025),
    hi_end=quantile(value, probs=0.975)) %>% 
  ggplot(aes(x=lambda, group=metric)) + 
    geom_line(aes(y=mean_INB)) + 
    geom_line(aes(y=lo_end), linetype="dashed") + 
    geom_line(aes(y=hi_end), linetype="dashed")

```

# Code for Exercises

## `Model_1_decision_model_odc.txt`

This is the code for Exercise 3.1, a deterministic decision tree comparing the use of prophylactic neuaminidase inhibitors (NIs) to standard care for influenza.

```{r model_1_decision_model}

model_code <- "model
{ 

  # DECISION MODEL
  costtrt <- ((1-p2) *(cdoc+cdrug) +p2* (cdoc+(cdrug)+cflu))
  costctl <- ((1-p1)*(cdoc)+p1*(cdoc+cflu))
  diff.cost <- costtrt-costctl

}" %>% textConnection

#DATA

data <- list(
  cdrug=118,    # cost of neuraminidase treatment
  cdoc=19,      # cost of visit to GP
  cflu=100,     # cost of treating flu
  p1=0.05,      # probability of contracting flu under standard care
  p2=0.03       # probability of contracting flu under NI prophylaxis				   
)

#RESULTS

results <- jags(
  data = data,
  # inits = NA,
  parameters.to.save = c("cdrug", "cdoc", "cflu", "p1", "p2",
                         "costtrt", "costctl", "diff.cost"),
  model.file = model_code,
  n.chains = 1,
  n.iter=1
)


results

```

# Errata

Text that should be deleted is marked in [square brackets]; text that should be added is marked in {curly braces}.

* p 47:
  - Equation 3.2 is about Odds Ratio, not Relative Risk. It should be
  $ var(ln(OR)) = 1/a + 1/b + 1/c + 1/d $
  
* p 54:
  - "estimate [the ]the payoffs"

* p 59:
  - An additional set of parentheses is needed to make the whole expression an argument of the log function: `theta <- log( (a/(a+b))/(c/(c+d)) )`

* p 60:
  - `dose` is hard-coded as `3`, where in the deterministic version it was specified as a data variable. It would be more consistent to do it the same way in both programs.
  
* p 61:
  - "However, in the situation where the new treatment is either more effective [but less]{and more} costly ..."

* p 63:
  - Figure 3.6(a) basically repeats Figure 3.4.
  - Figure 3.6(c) basically repeats Figure 3.5.

* p 69:
  - "... relevancy of evidence should be defined more broadly than afforded in Table 3.[1]{7}."
