---
title: 'Chapter 9: Mixed and indirect treatment comparisons'
author: "Bob"
date: "2025-03-01"
output: html_document
---

<font color="#008800">

These results need to be carefully checked against WinBUGS.

The first 3 examples use the `rank` function, which is parameterized differently in JAGS than in WinBUGS. See [this stackoverflow post](https://stackoverflow.com/questions/55628471/rank-function-in-jags).

  "The OpenBugs manual gives the definition as
  `rank(v, s) number of components of v less than or equal to s`
  which is what I had based my code on, but the manual is wrong and it should be
  `rank(v, s) number of components of v less than or equal to the s-th component of v`"


The fourth example has multi-dimensional data that I reformatted in code.
</font>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')
```


# `mtc_fe`
```
Fixed effect model for Hasselblad's data
==================================
24 trials (including two 3-arm-trials), 
50 data points,
4 treatment: 
          1=no contact (baseline)
          2=self help
          3=individual counselling
          4=group counselling
==================================
```


```{r mtc_fe}
# Model

model_code <- "
model{

  for(i in 1:50)  {                                           # loop over 50 observation
    r[i] ~ dbin(p[i],n[i])                                    # likelihood
    logit(p[i]) <- mu[s[i]]+ d[t[i]] - d[b[i]]                # model
    # Deviance residuals for data i                                                                                       
    rhat[i] <- p[i] * n[i]                                    # predicted values and deviance                           
    dev[i] <- 2 * (r[i] * (log(r[i])-log(rhat[i]))  +  (n[i]-r[i]) * (log(n[i]-r[i]) - log(n[i]-rhat[i])))    
  }
  sumdev <- sum(dev[])                                        # summed deviance

  for (j in 1:24) { mu[j] ~ dnorm(0,.0001)}                   # vague priors for baselines
  
  d[1] <- 0                                                   # set effect of A rel to A as 0
  for (k in 2:4) {d[k] ~ dnorm(0,.001) }                      # flat priors for 3 basic treatment parameters
  
  # Absolute treatment effects  
  precA <- pow(.38,-2)                                        # prior precision for Treatment A, sd=.38 
  A ~ dnorm(-2.6,precA)                                       # external info on A. 
  for (k in 1:4) { logit(T[k]) <- A + d[k] }                  # absolute effects T[k] for each treatment
                     
  # Rank the treatment effects (with 1=best) & record the best treatment
  rank_T_results <-  rank(T[]) # !!! Bob
  for (k in 1:4) { 
    # rk[k] <- 5 - rank(T[],k)
    rk[k] <- 5 - rank_T_results[k]
    best[k] <- equals(rk[k],1)
  }

  # All pairwise log odds ratios and odds ratios
  for (c in 1:3) { 
    for (k in (c+1):4) { 
      lor[c,k] <- d[k] - d[c]
      log(or[c,k]) <- lor[c,k]
    }
  }
}
" %>% textConnection


# Data: same as in random effects model above

data <- "s  t        r       n    b
1    1         9     140    1
1    3        23    140    1
1    4        10     138    1
2    2        11       78    2
2    3        12       85    2
2    4        29     170    2
3    1        75      731    1
3    3       363     714    1
4    1          2      106    1
4    3          9      205    1
5    1        58      549    1
5    3       237   1561    1
6    1          0        33    1
6    3         9         48    1
7    1         3       100    1
7    3        31        98    1
8     1        1          31    1
8     3       26        95    1
9    1         6         39    1
9    3       17         77    1
10   1      79      702    1
10   2      77      694    1
11   1      18      671    1
11   2       21     535     1
12   1       64     642    1
12   3      107    761    1 
13   1         5       62    1
13   3         8       90    1
14   1      20      234    1
14   3      34      237    1 
15   1        0        20     1
15   4        9        20    1
16   1        8     116    1
16    2     19     149    1 
17    1     95    1107    1
17    3    143   1031    1
18    1     15     187    1
18    3     36     504    1
19    1     78     584    1 
19    3     73     675    1 
20    1     69   1177    1 
20    3     54     888    1  
21    2     20      49    2
21    3     16      43    2
22    2       7      66    2
22     4    32    127    2
23     3    12      76    3
23     4    20      74    3
24     3      9       55    3
24     4      3       26    3
" %>% read.table(text=., header=TRUE) %>% as.list


# Starting/initial values: same as random effects model without sd
initial_values <- list(
  #initial 1
  list(
  d=c(NA,0,0,0),mu=c(0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0), A=-2
  ),
  
  #initial 2
  list(
  d=c(NA,0.1,-1,-0.2),mu=c(1,-1,-2,0,0,    -2,1,0,2,2,    1,-1,-2,0,0,   -2,1,0,2,2,   -2,-0.5,-3,0.5), A=-3
  )
  
)


parameters_to_save <- c("r", "delta", "tau") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

# plot(results)
```


# `mtc_re`

```
Random effect model for Hasselblad's data
==================================
24 trials (including two 3-arm-trials), 
50 data points,
4 treatment: 
          1=no contact (baseline)
          2=self help
          3=individual counselling
          4=group counselling
==================================
```

```{r mtc_re}


# Model

model_code <- "model{

  for(i in 1:50)  {                                         #  loop over 50 observation
    r[i] ~ dbin(p[i],n[i])                                  # likelihood
    logit(p[i]) <- mu[s[i]]+ delta[i]*(1-equals(t[i],b[i])) # model
    delta[i] ~ dnorm(md[i],tau)                             # Draw random effect from distribution
    md[i] <- d[t[i]] - d[b[i]]                              #  Mean of distribution
    # Deviance residuals for data i                                                                                       
    rhat[i] <- p[i] * n[i]                                  # predicted values and deviance                           
    dev[i] <- 2 * (r[i] * (log(r[i])-log(rhat[i]))  +  (n[i]-r[i]) * (log(n[i]-r[i]) - log(n[i]-rhat[i])))    
  }
  sumdev <- sum(dev[])                                      # summed deviance
  
  for (j in 1:24) { mu[j] ~ dnorm(0,.0001) }                # vague priors for baselines
                       
  d[1] <- 0                                                 # set effect of A rel to A as 0
  for (k in 2:4) {d[k] ~ dnorm(0,.001) }                    # flat priors for 3 basic treatment parameters
  
  tau <- pow(sd,-2)
  sd ~ dunif(0,2)
  
  # Absolute treatment effects  
  precA <- pow(.38,-2)                                      # prior precision for Treatment A, sd=.38 
  A ~ dnorm(-2.6,precA)                                     # external info on A. 
  for (k in 1:4) { logit(T[k]) <- A + d[k] }                # absolute effects T[k] for each treatment
                     
  #Rank the treatment effects (with 1=best) & record the best treatment
  rank_T_results <- rank(T[]) # !!! rank outside loop
  for (k in 1:4) { 
    # !!! rk[k] <- 5 - rank(T[],k)
    rk[k] <- 5 - rank_T_results[k]
    best[k] <- equals(rk[k],1)
  }

  #All pairwise log odds ratios and odds ratios
  for (c in 1:3) { 
    for (k in (c+1):4) { 
      lor[c,k] <- d[k] - d[c]
      log(or[c,k]) <- lor[c,k]
    }
  }

}
" %>% textConnection


# Data

data <- "s  t        r       n    b
1    1         9     140    1
1    3        23    140    1
1    4        10     138    1
2    2        11       78    2
2    3        12       85    2
2    4        29     170    2
3    1        75      731    1
3    3       363     714    1
4    1          2      106    1
4    3          9      205    1
5    1        58      549    1
5    3       237   1561    1
6    1          0        33    1
6    3         9         48    1
7    1         3       100    1
7    3        31        98    1
8     1        1          31    1
8     3       26        95    1
9    1         6         39    1
9    3       17         77    1
10   1      79      702    1
10   2      77      694    1
11   1      18      671    1
11   2       21     535     1
12   1       64     642    1
12   3      107    761    1 
13   1         5       62    1
13   3         8       90    1
14   1      20      234    1
14   3      34      237    1 
15   1        0        20     1
15   4        9        20    1
16   1        8     116    1
16    2     19     149    1 
17    1     95    1107    1
17    3    143   1031    1
18    1     15     187    1
18    3     36     504    1
19    1     78     584    1 
19    3     73     675    1 
20    1     69   1177    1 
20    3     54     888    1  
21    2     20      49    2
21    3     16      43    2
22    2       7      66    2
22     4    32    127    2
23     3    12      76    3
23     4    20      74    3
24     3      9       55    3
24     4      3       26    3
" %>% read.table(text=., header=TRUE) %>% as.list


# Starting/initial values
initial_values <- list(
  #initial 1
  list(
    d=c(NA,0,0,0),
    mu=c(0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0), 
    A=-2,
    sd=1
  ),
  
  #initial 2
  list(
    d=c(NA,0.1,-1,-0.2),
    mu=c(1,-1,-2,0,0,    -2,1,0,2,2,    1,-1,-2,0,0,   -2,1,0,2,2,   -2,-0.5,-3,0.5), 
    A=-3,
    sd=.5
  )

  
)


parameters_to_save <- c("r", "delta", "tau") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

# plot(results)
```


# `mtc_re_ceac`

```
Random effects model for Hasselblad's data
==================================
24 trials (including two 3-arm-trials), 
50 data points,
4 treatment: 
          1=no contact (baseline)
          2=self help
          3=individual counselling
          4=group counselling
==================================
```
<font color="#008800">
The rank function call here is complicated because it is doubly nested. These results need to be checked against WinBUGS with particular care, in case I did this wrong.
</font>


```{r mtc_re_ceac}


# Model

model_code <- "model{

# Model for log-odds of smoking cessation, for three types of trial indicated by b[i]
  for(i in 1:50){ 
    r[i] ~ dbin(p[i],n[i])
    logit(p[i])<-mu[s[i]]+ delta[i]*(1-equals(t[i],b[i]))
    #Random effects model for log-odds ratios
    delta[i] ~ dnorm(md[i],prec)
    md[i] <- d[t[i]] - d[b[i]]
    #Deviance residuals for data i                                                                                       
    rhat[i] <- p[i] * n[i]                                                                                                          
    dev[i] <- 2 * (r[i] * (log(r[i])-log(rhat[i]))  +  (n[i]-r[i]) * (log(n[i]-r[i]) - log(n[i]-rhat[i])))    
  }
		
  sumdev <- sum(dev[])

  #Fixed effect priors
  for(j in 1:24){ 
    mu[j]~dnorm(0,.0001)
  }
  prec <- 1/(sd*sd)
  sd ~ dunif(0,2)
  
  #Define the baseline shifts, and give priors for log-odds ratios
	d[1]<-0
	for (k in 2:4){
	  d[k] ~ dnorm(0,.001) 
  }


  # Absolute treatment effects  
  precA <- pow(.38,-2)                                         # prior precision for Treatment A, sd=.38 
  A ~ dnorm(-2.6,precA)                                        # external info on A. 
  for (k in 1:4) { logit(T[k]) <- A + d[k] }                    # absolute effects T[k] for each treatment
  
  # Rank the treatment effects (with 1=best) & record the best treatment
  rank_T_results <- rank(T[]) # !!! call rank function outside of loop
  for (k in 1:4){
    # !!! rk[k]<- 5 - rank(T[],k)
    rk[k] <- 5 - rank_T_results[k]
    best[k] <- equals(rk[k],1)
  }
  
  # All pairwise log odds ratios and odds ratios
  for (c in 1:3){
  	for (k in (c+1):4){
  		lor[c,k] <- d[k] - d[c]
  		log(or[c,k]) <- lor[c,k]
  	}
  }
  
  #Economic data
  #Life-years gained
  lyg~dnorm(15,.0625)
  cost[1]<-0
  cost[2]<-200
  cost[3]<-6000
  cost[4]<-600
  
  # !!! As above, the call to `rank` needs to be moved out of the inner loop, then we index into the result.
  # !!! Here it is more complicated because it needs to be done inside a nested loop.
  # !!! I broke it into two passes: first compute nb, then go back and handle the rank calls to compute p.ce.
  
  # for (w in 1:50){
  # 	for (k in 1:4){
  # 		nb[k,w] <- T[k]*lyg*w*1000 - cost[k]
  # 		p.ce[k,w] <- equals(rank(nb[,w],k),4)
  # 	}	
  # }
  
  for (w in 1:50){
  	for (k in 1:4){
  		nb[k,w] <- T[k]*lyg*w*1000 - cost[k]
  	}
  }
  
  for (w in 1:50){
    rank_w_results[1:4,w] <- rank(nb[,w])
  	for (k in 1:4){
  		p.ce[k,w] <- equals(rank_w_results[k,w],4)
  	}
  }

}


" %>% textConnection


# Data: same as in random effects model above

data <- "s  t        r       n    b
1    1         9     140    1
1    3        23    140    1
1    4        10     138    1
2    2        11       78    2
2    3        12       85    2
2    4        29     170    2
3    1        75      731    1
3    3       363     714    1
4    1          2      106    1
4    3          9      205    1
5    1        58      549    1
5    3       237   1561    1
6    1          0        33    1
6    3         9         48    1
7    1         3       100    1
7    3        31        98    1
8     1        1          31    1
8     3       26        95    1
9    1         6         39    1
9    3       17         77    1
10   1      79      702    1
10   2      77      694    1
11   1      18      671    1
11   2       21     535     1
12   1       64     642    1
12   3      107    761    1 
13   1         5       62    1
13   3         8       90    1
14   1      20      234    1
14   3      34      237    1 
15   1        0        20     1
15   4        9        20    1
16   1        8     116    1
16    2     19     149    1 
17    1     95    1107    1
17    3    143   1031    1
18    1     15     187    1
18    3     36     504    1
19    1     78     584    1 
19    3     73     675    1 
20    1     69   1177    1 
20    3     54     888    1  
21    2     20      49    2
21    3     16      43    2
22    2       7      66    2
22     4    32    127    2
23     3    12      76    3
23     4    20      74    3
24     3      9       55    3
24     4      3       26    3
" %>% read.table(text=., header=TRUE) %>% as.list


# Starting/initial values: same as random effects model without sd
initial_values <- list(
  #initial 1
  list(
  d=c(NA,0,0,0),sd=1,mu=c(0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0)
  ),
  
  #initial 2
  list(
  d=c(NA,0.1,-1,-0.2),sd=.5,mu=c(1,-1,-2,0,0,    -2,1,0,2,2,    1,-1,-2,0,0,   -2,1,0,2,2,   -2,-0.5,-3,0.5)
  )
  
)


parameters_to_save <- c("cost") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

# plot(results)
```


# mtc_re_multiarm

Random effects model for multi-arm trials (any number of arms).

<font color="#008800"?
NOTE: This example has 3-dimensional data consisting of three matrixes ('r', 'n', and 't') and a vector 'na'; I re-format them into a list of smaller data structures.
</font>


```{r mtc_re_multiarm}

model_code <- "
model{
  for(i in 1:NS){ 
    w[i,1] <- 0
    delta[i,t[i,1]]<-0
    mu[i] ~ dnorm(0,.0001)                                    # vague priors for 24 trial baselines
    for (k in 1:na[i])  { 
      r[i,k] ~ dbin(p[i, t[i,k]], n[i,k])                     # binomial likelihood
      logit(p[i, t[i,k]])<-mu[i] + delta[i,t[i,k]] 
    }                                                         # model
    for (k in 2:na[i]) {
      delta[i,t[i,k]] ~ dnorm(md[i, t[i,k]], taud[i,t[i,k]])  # trial-specific LOR distributions
      md[i,t[i,k]] <-  d[t[i,k]] - d[t[i,1]]  + sw[i,k]       # mean of LOR distributions
      taud[i,t[i,k]] <- tau *2*(k-1)/k                        # precision of LOR distributions
      w[i,k] <- (delta[i,t[i,k]]  - d[t[i,k]] + d[t[i,1]])    # adjustment, multi-arm RCTs
      sw[i,k] <- sum(w[i,1:(k-1)])/(k-1)                      # cumulative adjustment for multi-arm trials 
            # !!! added parentheses around k-1 in 1:(k-1)
    }
  }
  
  d[1] <- 0
  for (k in 2:NT){ d[k] ~ dnorm(0,.0001) }            #  vague priors for basic parameters
  
  sd ~ dunif(0,2)                                     #  vague prior for random effects standard deviation 
  tau <- 1 / pow(sd,2)
}
" %>% textConnection

# "r[,1]  n[,1]	r[,2]	n[,2]	r[,3]	n[,3]	t[,1]  t[,2] t[,3]   na[]

data_df <- "r1  n1	r2	n2	r3	n3	t1  t2 t3   na   
  9	  140	  23	  140	 10	138	 1      3     4      3
 11	   78	 12	    85	   29    170	2      3     4      3
 75	 731	363	 714	   NA	  1	 1      3   NA     2 
  2	  106	  9	   205	   NA	  1	 1      3   NA     2
 58	 549	237	1561	 NA	   1     1      3   NA     2
  0	    33	  9	     48	   NA	  1	 1      3   NA     2
  3	  100	 31	    98	   NA	  1	 1      3   NA     2 
  1	    31	 26	    95	   NA	  1	 1      3   NA     2
  6	    39	 17	    77	   NA     1	 1      3   NA     2
79	  702	 77	 694	   NA	   1     1      2   NA     2 
18	  671	 21	 535	   NA	   1 	1      2   NA     2
64	  642   107	 761	   NA	   1	 1      3   NA     2
  5	   62	   8	    90	   NA	  1	 1      3   NA     2
20	 234	 34	  237	   NA	   1	 1      3   NA     2
  0	  20	    9	     20      NA	   1	 1      4   NA     2
  8	 116	  19	  149	  NA	   1	 1      2   NA     2
95	1107  143   1031 	   NA	   1	 1      3   NA      2
15	 187	  36	504 	   NA	   1	 1      3   NA      2
78	 584	 73	675  	   NA       1	 1      3   NA      2
69    1177	 54	888	    NA	   1	 1     3   NA       2
20	    49	 16	  43	    NA	    1	 2     3   NA      2
  7	    66	  32   127	    NA	    1	 2     4   NA      2
12	  76	   20	  74        NA	    1	 3     4   NA      2
  9	   55	    3	   26  	  NA	   1	 3     4   NA     2" %>% read.table(text=., header=TRUE)

# more_data <- list(NT=4, NS=24)

data <- with(data_df, list(
  r = cbind(r1, r2, r3),
  n = cbind(n1, n2, n3),
  t = cbind(t1, t2, t3),
  na = na,
  NT=4, 
  NS=24
))

initial <- list(
  list(
    d=c(NA,0,0,0),
    sd=1,
    mu=c(0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0,0,  0,0,0,0)
  ),
  
  list(
    d=c(NA,0.1,-1,-0.2),
    sd=.2,
    mu=c(1,-1,-2,0,0,    -2,1,0,2,2,    1,-1,-2,0,0,   -2,1,0,2,2,   -2,-0.5,-3,0.5)
  )
  
)


parameters_to_save <- c("tau", "sd", "delta") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)
```