---
title: 'Chapter 4: Meta-analysis Using Bayesian Methods'
author: "Bob"
date: "2025-02-26"
output: html_document
---

# TO DO: 
* Models `Cholesterol_prior_sens` and `asprin_logistic_random_1_priors_sensitivity` fail with `Attempt to redefine node` errors. These chunks are set to `eval=FALSE`.
* Describe the point of each model.
* Be sure appropriate outputs are summarized or plotted. What other plots are relevant?
* Clean up code: re-use datasets as appropriate, smarter table loading (compute k from the length of the dataframe; use this value to generate initial values), ... 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, warn=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')
```

# Aspirin

## `asprin_normal_fixed_1.txt`

This is the model from Figure 4.1 (with a few extra digits of precision in the data).

```{r asprin_normal_fixed_1}

model_code <- "model 
{ 
  for (i in 1:Nstud)
    {
      P[i] <- 1/V[i]
      y[i] ~ dnorm(d, P[i])
    }
	d ~ dnorm(0, 1.0E-5)
	OR <- exp(d)
}" %>% textConnection

# DATA
data <- list(
  y=c(.3289011, .3845458, .2195622, .2222206, .2254672, -.1246363, .1109658), 
  V=c(.0388957,.0411673,.0204915,.0647646,.0351996,.0096167,.0015062), 
  Nstud=7
)

# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(d=0)
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```


## `asprin_normal_random_1.txt`

This is the model from Figure 4.3, except it uses lower case for the y[] variable vector.

```{r asprin_normal_random_1}

model_code <- "model 
{ 
  for (i in 1:Nstud)
    {
      P[i] <- 1/V[i]
      y[i] ~ dnorm(delta[i], P[i])
      delta[i] ~ dnorm(d, prec) 
    }	
	d ~ dnorm(0, 1.0E-5)
	OR <- exp(d)
	tau ~ dunif(0,10)
	tau.sq <- tau*tau
	prec <- 1/(tau.sq)
}" %>% textConnection

# DATA  (the same as for the fixed effects model above)
data <- list(
  y=c(.3289011, .3845458, .2195622, .2222206, .2254672, -.1246363, .1109658), 
  V=c(.0388957,.0411673,.0204915,.0647646,.0351996,.0096167,.0015062), 
  Nstud=7
)

# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(d = 0, tau = 1, delta = c(0,0,0,0,0,0,0))
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d", "prec", "tau.sq"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```

## `asprin_logistic_random_1.txt`

This is the model from Figure 4.5 (except that the line `OR <- exp(d)` is in a different place).

```{r asprin_logistic_random_1}

model_code <- "model
{
  for( i in 1 : Nstud ) {
    rA[i] ~ dbin(pA[i], nA[i])
    rB[i] ~ dbin(pB[i], nB[i])
    logit(pA[i]) <- mu[i]
    logit(pB[i]) <- mu[i] + delta[i]
    mu[i] ~ dnorm(0.0,1.0E-5)
    delta[i] ~ dnorm(d, prec)
  }
  OR <- exp(d)   # !!! this line comes last in Figure 4.5
  d ~ dnorm(0.0,1.0E-6)
  tau ~ dunif(0,10)
  tau.sq <- tau*tau
  prec <- 1/(tau.sq)
}" %>% textConnection

# DATA
data <- list(
  rA=c(49,44,102,32,85,246,1570), 
  rB=c(67,64,126,38,52,219,1720), 
  nA=c(615,758,832,317,810,2267,8587), 
  nB=c(624,771,850,309,406,2257,8600), 
  Nstud=7
)


# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(d=0, tau=1, delta=c(0,0,0,0,0,0,0), mu=c(0,0,0,0,0,0,0))
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d", "prec", "tau.sq"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```

## `asprin_logistic_random1_predict.txt`

```{r asprin_logistic_random1_predict}

model_code <- "model
{
  for( i in 1 : Nstud ) {
    rA[i] ~ dbin(pA[i], nA[i])
    rB[i] ~ dbin(pB[i], nB[i])
    logit(pA[i]) <- mu[i]
    logit(pB[i]) <- mu[i] + delta[i]
    mu[i] ~ dnorm(0.0,1.0E-5)
    delta[i] ~ dnorm(d, prec)
  }
  
  OR <- exp(d)
  d ~ dnorm(0.0,1.0E-6)
  tau~dunif(0,10)
  tau.sq<-tau*tau
  prec<-1/(tau.sq)
  
  d.new ~ dnorm(d, prec)
  OR.new <-exp(d.new)
}" %>% textConnection

# DATA
data <- list(
  rA=c(49,44,102,32,85,246,1570), 
  rB=c(67,64,126,38,52,219,1720), 
  nA=c(615,758,832,317,810,2267,8587), 
  nB=c(624,771,850,309,406,2257,8600), 
  Nstud=7
)


# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(d=0, tau =1, delta=c(0,0,0,0,0,0,0), mu=c(0,0,0,0,0,0,0))
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "OR.new", "d.new", "delta", "tau.sq"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```

## `asprin_logistic_random_1_priors_sensitivity`

```
RUNTIME ERROR:
Compilation error on line 13.
Attempt to redefine node rBx[1,1]
```

```{r asprin_logistic_random_1_priors_sensitivity, eval=FALSE}


model_code <- "
model
	{
	
  ## replicating the data 2 times to fit to 3 models in total where j indexes alternative prior distributions
  
  for (j in 1:3)
    {
    
      for( i in 1 : Nstud )
      {
        rAx[j,i] ~ dbin(pA[j,i], nA[i])
        rAx[j,i] <- rA[i]
        rBx[j,i] ~ dbin(pB[j,i], nB[i])
        rBx[j,i] <- rB[i]
        logit(pA[j,i]) <- mu[j,i]
        logit(pB[j,i]) <- mu[j,i] + delta[j,i]
        mu[j,i] ~  dnorm(0.0,1.0E-5)
        delta[j,i] ~ dnorm(d[j], prec[j])
      }
    }
  
  for (j in 1:3)
    {
      OR[j] <- exp(d[j])
      d[j] ~ dnorm(0.0,1.0E-6)
      m[j]~dnorm(0.0,1.0E-6)
    }

	
  # Setting three different priors for the between study variance
  
  prec[1] <- 1/tau.sq[1]
  tau.sq[1] <- tau[1]*tau[1]
  tau[1] ~ dnorm(0,1.0E-6)I(0,)
  
  prec[2] ~ dgamma(0.001,0.001)
  tau.sq[2] <- 1/prec[2]
  tau[2] <- sqrt(tau.sq[2])
  
  tau[3] ~ dunif(0,10)
  tau.sq[3] <- tau[3]*tau[3]
  prec[3] <- 1/tau.sq[3]
	
	}
" %>% textConnection

# DATA
data <- list(
  rA=c(49,44,102,32,85,246,1570), 
  rB=c(67,64,126,38,52,219,1720), 
  nA=c(615,758,832,317,810,2267,8587), 
  nB=c(624,771,850,309,406,2257,8600), 
  Nstud=7
)


# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  # WinBUGS code has extra closing parentheses (???)
  list(
    prec=c(NA,0.1,NA),  
    tau=c(0.01, NA, 0.01), 
    mu = structure(
      .Data=c(0,0,0,0,0,0,0,  0,0,0,0,0,0,0,  0,0,0,0,0,0,0),
      .Dim=c(3,7)),
    delta = structure(
      .Data=c(0,0,0,0,0,0,0,   0,0,0,0,0,0,0,   0,0,0,0,0,0,0),
      .Dim=c(3,7)), 
    d =c(0.1,0.1,0.1)
  )

)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d", "delta", "tau.sq", "rAx"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```


# Cholesterol

## `chol_model.txt`, `chol_data.txt`

```
Error parsing model file:
syntax error on line 5 near "var"
```

Maybe `var` is a reserved word or something? I changed it to `V` (as in the next model), which seems to work.

```{r chol_model.txt}

model_code <- "model 
{ 
  for (i in 1:k)
  {
    precision[i] <- 1/V[i]
    logor[i] ~ dnorm(theta, precision[i])
  }
	theta ~ dnorm(0, 1.0E-5)
	OR <- exp(theta)
}
" %>% textConnection

# DATA
data <- list(k=34)

data_df <- "logor	V
0.4893853	0.1987757
-0.0683811	0.054427
-0.4876372	0.0730886
-0.1630902	0.1108992
1.469676	1.186739
0.0057637	0.0608936
-0.1496961	0.0799553
-1.94591	2.351288
-0.0470209	0.0194635
-0.3824268	0.0552552
-0.5508629	0.0615112
-0.0984401	0.0510244
0.2939126	0.1708592
-0.4883528	1.582492
-0.4354064	0.0030637
-0.2885568	0.0496847
0.4696847	0.0716345
0.2692525	0.0101539
-1.563976	0.8819983
-0.3517397	0.3733933
-0.0480808	0.0298812
-1.109251	2.687944
0.0129744	0.0469846
-0.4124823	0.0383304
-0.0233855	0.0200697
0.0803459	0.0082181
-0.2820902	0.0420423
-1.098612	2.707904
0.521464	2.696409
-0.2501722	0.477032
1.025401	0.3643915
-0.7528253	0.0676265
-1.580711	2.272708
0.5030903	0.1426682" %>% read.delim(text=., sep="\t")

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}



results <- jags(
  data = data,
  # inits = initial_values,
  parameters.to.save = c("OR", "theta"),
  model.file = model_code,
  n.chains = 1, # length(initial_values),
  # n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000 #, n.thin = 2
)

summary(results)

plot(results)

```

# `Cholesterol_fixed.txt`

This is the file for Model 4.1 from the Exercises, referenced on p. 89.

Model 1: Generic fixed effect model applied to the cholesterol lowering dataset

```{r Cholesterol_fixed}

model_code <- "model 
{ 
  for (i in 1:k)
  {
    P[i] <- 1/V[i]
    logor[i] ~ dnorm(d, P[i])
  }
	d ~ dnorm(0, 1.0E-5)
	OR <- exp(d)
}" %>% textConnection

# DATA: re-used from previous model. Here the name of the second column in the data table is `V` instead of `var`. I changed that in the previous model because the name `var` seemed to be causing problems.

# data <- list(k=34)
# 
# data_df <- "logor	V
# 0.4893853	0.1987757
# -0.0683811	0.054427
# -0.4876372	0.0730886
# -0.1630902	0.1108992
# 1.469676	1.186739
# 0.0057637	0.0608936
# -0.1496961	0.0799553
# -1.94591	2.351288
# -0.0470209	0.0194635
# -0.3824268	0.0552552
# -0.5508629	0.0615112
# -0.0984401	0.0510244
# 0.2939126	0.1708592
# -0.4883528	1.582492
# -0.4354064	0.0030637
# -0.2885568	0.0496847
# 0.4696847	0.0716345
# 0.2692525	0.0101539
# -1.563976	0.8819983
# -0.3517397	0.3733933
# -0.0480808	0.0298812
# -1.109251	2.687944
# 0.0129744	0.0469846
# -0.4124823	0.0383304
# -0.0233855	0.0200697
# 0.0803459	0.0082181
# -0.2820902	0.0420423
# -1.098612	2.707904
# 0.521464	2.696409
# -0.2501722	0.477032
# 1.025401	0.3643915
# -0.7528253	0.0676265
# -1.580711	2.272708
# 0.5030903	0.1426682" %>% read.delim(text=., sep="\t")
# 
# for (colname in names(data_df)){
#   data[[colname]] <- data_df[[colname]]
# }

# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(d=0)
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

plot(results)

```

## `Cholesterol_random.txt`

This is the file for Model 4.2 from the Exercises, referenced on p. 90.

Model 2: Generic random effect model applied to the cholesterol lowering dataset

```{r Cholesterol_random}


model_code <- "model 
{ 
  for (i in 1:k)
    {
      P[i] <- 1/V[i]
      logor[i] ~ dnorm(delta[i], P[i])
      delta[i] ~ dnorm(d, prec) 
    }	
	d ~ dnorm(0, 1.0E-5)
	OR <- exp(d)
	
	tau ~ dunif(0,10)
	tau.sq <- tau*tau
	prec <- 1/(tau.sq)
	d.new ~ dnorm(d, prec)
}" %>% textConnection

# DATA: re-used from previous chunks.


initial_values <- list(
  list(d=0, 
       delta=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0), 
       d.new=0)
)

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "d"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000,
  n.thin = 2
)

summary(results)

```

## `Cholesterol_random_logistic.txt`

This is the file for Model 4.3 in the Exercises, p90.

```{r Cholesterol_random_logistic}
model_code <- "model
	{
	   for( i in 1 : k ) {
	   	rA[i] ~ dbin(pA[i], nA[i])
		   rB[i] ~ dbin(pB[i], nB[i])
		logit(pA[i]) <- mu[i]
		logit(pB[i]) <- mu[i] + delta[i]
		mu[i] ~ dnorm(0.0,1.0E-5)
		delta[i] ~ dnorm(d, prec)
	   }
		OR <- exp(d)
	    d ~ dnorm(0.0,1.0E-6)
	
		tau~dunif(0,10)
		tau.sq<-tau*tau
		prec<-1/(tau.sq)
	}

" %>% textConnection

# DATA
data <- list(k=34)

data_df <- "nB	nA	rB	rA
50	50	17	12
285	147	70	38
156	119	37	40
123	129	20	24
54	26	8	1
427	143	81	27
199	194	28	31
30	33	0	3
424	422	174	178
206	206	41	55
244	253	31	51
350	367	42	48
47	48	23	20
23	29	1	2
5552	2789	1025	723
1149	1129	37	48
221	237	39	28
5331	5296	236	181
88	30	2	3
71	72	5	7
1906	1900	68	71
94	94	0	1
2051	2030	44	43
279	276	61	82
1018	1015	111	113
4541	4516	269	248
421	417	49	62
48	49	0	1
94	52	1	0
79	78	4	5
6582	1663	33	3
204	202	28	51
30	60	0	4
311	317	19	12" %>% read.delim(text=., sep="\t")

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}

initial_values = list(
  list(d=0, 
       tau=1, 
       delta=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0), # rep(0, data$k)
       mu=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))
  )

results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "theta"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000 #, n.thin = 2
)

summary(results)

plot(results)

```

## `Cholesterol_random_logistic_2`

Also referenced on p90.

Model 3 (Cont) Question 5: Direct OR model applied to the cholesterol lowering dataset

```{r Cholesterol_random_logistic_2}
model_code <- "

model
{
  for( i in 1 : k ) {
    rA[i] ~ dbin(pA[i], nA[i])
    rB[i] ~ dbin(pB[i], nB[i])
    logit(pA[i]) <- mu[i]
    logit(pB[i]) <- mu[i] + delta[i]
    mu[i] ~ dnorm(0.0,1.0E-5)
    delta[i] ~ dnorm(d, prec)
  }
  OR <- exp(d)
  d ~ dnorm(0.0,1.0E-6)
  
  tau~dunif(0,10)
  tau.sq<-tau*tau
  prec<-1/(tau.sq)
  
  # Storing theta.pooled in delta[37] to add to caterpillar plot
  delta[37] <- d
  
  # Replicate from a predictive distribution for a future study
  delta[39] ~ dnorm(d, prec)
  
}
" %>% textConnection

# DATA
data <- list(k=34)

data_df <- "nB	nA	rB	rA
50	50	17	12
285	147	70	38
156	119	37	40
123	129	20	24
54	26	8	1
427	143	81	27
199	194	28	31
30	33	0	3
424	422	174	178
206	206	41	55
244	253	31	51
350	367	42	48
47	48	23	20
23	29	1	2
5552	2789	1025	723
1149	1129	37	48
221	237	39	28
5331	5296	236	181
88	30	2	3
71	72	5	7
1906	1900	68	71
94	94	0	1
2051	2030	44	43
279	276	61	82
1018	1015	111	113
4541	4516	269	248
421	417	49	62
48	49	0	1
94	52	1	0
79	78	4	5
6582	1663	33	3
204	202	28	51
30	60	0	4
311	317	19	12" %>% read.delim(text=., sep="\t")

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}

initial_values = list(
  list(
    d=0, 
    tau=2, 
    delta=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,NA,NA,NA,NA,1),  
    mu=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0))

)


results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("OR", "theta"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000 #, n.thin = 2
)

summary(results)

plot(results)

```

## `Cholesterol_prior_sens`

This file is referenced for Model 4.4, p91.

Model 4: Direct OR model with built in sensitivity analysis to prior

```
RUNTIME ERROR:
Compilation error on line 14.
Attempt to redefine node rBx[1,1]
```

```{r Cholesterol_prior_sens, eval=FALSE}
model_code <- "
model
{
	
  ## replicating the data 2 times to fit to 3 models in total where j indexes alternative prior distributions

  for (j in 1:3)
  {
     
    for( i in 1 : k )
    {
      rAx[j,i] ~ dbin(pA[j,i], nA[i])
      rAx[j,i] <- rA[i]
      rBx[j,i] ~ dbin(pB[j,i], nB[i])
      rBx[j,i] <- rB[i]
      logit(pA[j,i]) <- mu[j,i]
      logit(pB[j,i]) <- mu[j,i] + delta[j,i]
      mu[j,i] ~ dnorm(0.0,1.0E-5)
      delta[j,i] ~ dnorm(d[j], prec[j])
    }
  }

  for (j in 1:3)
  {
    OR[j] <- exp(d[j])
    d[j] ~ dnorm(0.0,1.0E-6)
    m[j] ~ dnorm(0.0,1.0E-6)
  }

	
  # Setting three different priors for the between study variance
  
  prec[1] <-1/tau.sq[1]
  tau.sq[1] <-tau[1]*tau[1]
  tau[1] ~ dnorm(0,1.0E-6)I(0,)
  
  prec[2] ~ dgamma(0.001,0.001)
  tau.sq[2] <- 1/prec[2]
  tau[2] <- sqrt(tau.sq[2])
  
  tau[3] ~ dunif(0,50)
  tau.sq[3] <- tau[3]*tau[3]
  prec[3] <- 1/tau.sq[3]
	
}" %>% textConnection

# DATA
data <- list(k=34)

data_df <- "nB	nA	rB	rA
50	50	17	12
285	147	70	38
156	119	37	40
123	129	20	24
54	26	8	1
427	143	81	27
199	194	28	31
30	33	0	3
424	422	174	178
206	206	41	55
244	253	31	51
350	367	42	48
47	48	23	20
23	29	1	2
5552	2789	1025	723
1149	1129	37	48
221	237	39	28
5331	5296	236	181
88	30	2	3
71	72	5	7
1906	1900	68	71
94	94	0	1
2051	2030	44	43
279	276	61	82
1018	1015	111	113
4541	4516	269	248
421	417	49	62
48	49	0	1
94	52	1	0
79	78	4	5
6582	1663	33	3
204	202	28	51
30	60	0	4
311	317	19	12" %>% read.delim(text=., sep="\t")

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}

initial_values = list(

  list(
    prec=c(NA,0.1,NA),  
    tau=c(0.5, NA, 0.5), 
    mu = structure(.Data=c(0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0,
0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0,
0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0),.Dim=c(3,34)),

    delta = structure(.Data=c(
0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0,
0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0,
0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0,   0,0,0,0),.Dim=c(3,34)) #)
    , 
    d =c(0.1,0.1,0.1) 
  )

)


results <- jags(
  data = data,
  inits = initial_values,
  parameters.to.save = c("deviance", "theta"),
  model.file = model_code,
  n.chains = length(initial_values),
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000 #, n.thin = 2
)

summary(results)

plot(results)

```