---
title: 'Chapter 3: Introduction to Decision Models'
author: "Bob"
date: "2025-02-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, warn=FALSE, message=FALSE}
library('jagsUI')     # wrapper for rjags
library('mcmcplots')  # caterplot
library('coda')
library('dplyr')
```


# `caesarean_section_example.odc`

This is a copy of the caesarian section code from Chapter 7. Chapter 3 was supposed to be a simpler version, but the file in the supplementary materials archive appears to be exactly the same.

# To Do: break this example into two parts, meta-analysis and decision tree.

```{r caesarean_section_example_odc}

# Model

model_code <- "model
{

  # META-ANALYSIS ADJUSTING FOR BASELINE RISK

  for (i in 1:Num) 
  {
  	rc[i] ~ dbin(pc[i], nc[i])
  	rt[i] ~ dbin(pt[i], nt[i])	
  	logit(pc[i]) <- mu[i]
  	logit(pt[i]) <- mu[i] + delta[i] 
  	delta[i] <- delta.star[i] + beta * (mu[i]-(-1.09))
  	mu[i] ~ dnorm(0.0,1.0E-3)
  	delta.star[i] ~ dnorm(d, prec)
  }	
  
  d ~ dnorm(0.0,1.0E-6)
  OR <- exp(d)	
  d.pred ~ dnorm(d,prec)
  beta ~ dnorm(0.0,1.0E-6)
  
  tau ~ dunif(0,10)
  tau.sq <- tau*tau
  prec <- 1/tau.sq

  # DEVIANCE CONTRIBUTION

  for (i in 1:Num)
  {
    rthat[i] <- pt[i] * nt[i]
    dev[i] <- 2 * (rt[i] * (log(rt[i])-log(rthat[i]))  +  (nt[i]-rt[i]) * (log(nt[i]-rt[i]) - log(nt[i]-rthat[i])))
  }

  sumdev <- sum(dev[])
  
  # DATA FOR DECISION MODEL	
  
  # Probability of wound infection without antibiotics
  rc1 ~ dbin(p1,nc1)
  p1 ~ dbeta(1,1)
  
  # Probability of wound infection with antibiotics
  logit(p2) <- logit(p1) + d + beta*(logit(p1)-(-1.09))
  
  # Length of stay in hospital with wound infection
  loswd ~ dnorm(mnloswd,precwd)  
  precwd <- 1/pow(sdloswd/sqrt(numwd),2)		
  # Length of stay in hospital without wound infection	
  losnwd ~ dnorm(mnlosnwd,precnwd)  
  precnwd <- 1/pow(sdlosnwd/sqrt(numnwd),2)	
  
  # Antibiotics administration costs	
  drugadmin ~ dunif(4,10)
  
  # QALYs - wound infection
  QALYs.wd <- ((QALYwd/365)*loswd)+((Fullhealth/365)*(fllwupdays-loswd))
  # QALYs - no wound infection
  QALYs.nwd <- ((QALYnwd/365)*losnwd)+((Fullhealth/365)*(fllwupdays-losnwd))
  
  
  # DECISION MODEL	
  
  # Cost with prophylactic antibiotics									
  cst.trt <- (1-p2) * ((antibiotic+drugadmin) * 3 + (losnwd*inptnwd)) + p2 * ((antibiotic+drugadmin) * 3 + (loswd * inptwd))
  # Cost without prophylactic antibiotics
  cst.ctl <- (1-p1) * (losnwd*inptnwd) + p1 * (loswd*inptwd)		
  # Difference in cost		
  diff.cost <- cst.trt - cst.ctl
  
  # Number of wound infections avoided using prophylactic antibiotics
  diff.wd <- (nc1 * (p1 - p2))
  
  # QALYs - with prophylactic antibiotics	
  QALYs.trt <- (1-p2) * QALYs.nwd+p2 * QALYs.wd
  # QALYs - without prophylactic antibiotics	
  QALYs.ctl <- (1-p1) * QALYs.nwd+p1 * QALYs.wd
  # Difference in QALYs	
  diff.QALYs <- (QALYs.trt - QALYs.ctl)
  
  # Probability using prophylactic antibiotics costs less than not using prophylactic antibiotics
  Q <- step(cst.ctl - cst.trt)
  
  # Cost effectiveness ratio
  ce.ratio <- diff.cost/diff.wd	
  cu.ratio <- diff.cost/diff.QALYs			
  
  # INCREMENTAL NET BENEFIT

  for (k in 1:M) 
  {
  	Rc1[k] <- (k-1) * 2000
  	INB.QALYs[k] <- Rc1[k] * diff.QALYs - diff.cost		
  	ProbCU[k] <- step(INB.QALYs[k])
  }
}
" %>% textConnection


# Data

data <- list(
  Num=35, 
  rc1=41, 
  nc1=486, 
  inptwd=163.03, 
  inptnwd=107.26, 
  mnloswd=8.8, 
  sdloswd=3.5, 
  mnlosnwd=6.7,  
  sdlosnwd=7.1, 
  numwd=41, 
  numnwd=445, 
  antibiotic=5.67, 
  M=21, 
  QALYwd=0.68, 
  QALYnwd=0.88, 
  Fullhealth=1, 
  fllwupdays=20
)

data_df <- "rt	nt	rc	nc
4	133	28	136
2	68	1	56
5	55	8	55
3	100	0	33
4	183	3	44
0	46	4	55
0	119	1	39
2	60	6	30
1	46	1	49
0	50	2	50
1	43	1	47
0	80	3	80
2	196	14	190
2	124	2	58
2	74	8	78
4	167	5	140
0	48	2	49
0	102	1	99
0	85	3	43
6	136	16	130
0	74	2	74
2	61	2	61
3	146	9	132
1	64	7	65
0	16	1	16
2	31	6	37
1	50	2	49
2	26	0	27
3	133	12	66
6	244	17	117
1	52	2	61
2	43	4	50
2	48	3	45
3	40	1	40
1	50	4	50" %>% read.delim(text=., sep='\t')

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}

# Starting/initial values

initial_values <- list(
  list(
    tau=1,
    delta.star=c(0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0, 0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0),
    d=0,
    mu=c(0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0, 0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0),
    p1=1e-6, # 1e-6 works, 0.0 does not
    beta=1e-6  # 1e-6 works, 0.0 does not
  ),
  
  list(
    tau=0.1,
    delta.star=c(1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1),
    d=1, 
    mu = c(1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1), 
    p1=0.5, 
    beta=1)
)

parameters_to_save <- c("ProbCU", "INB.QALYs", "Rc1", "diff.QALYs","diff.cost","OR",
                        "beta","ce.ratio","cu.ratio","p1","p2",
                        # "sumdev",
                        "tau.sq")

results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

results_summary <- summary(results) %>% as.data.frame()

results_summary

```

# Code for Exercises

## `Model_1_decision_model_odc.txt`

This is the code for Exercise 3.1, a deterministic decision tree comparing the use of prophylactic neuaminidase inhibitors (NIs) to standard care for influenza.

```{r model_1_decision_model}

model_code <- "model
{ 

  #DECISION MODEL
  costtrt <- ((1-p2) *(cdoc+cdrug) +p2* (cdoc+(cdrug)+cflu))
  costctl <- ((1-p1)*(cdoc)+p1*(cdoc+cflu))
  diff.cost <- costtrt-costctl

}" %>% textConnection

#DATA

data <- list(
  cdrug=118,    # cost of neuraminidase treatment
  cdoc=19,      # cost of visit to GP
  cflu=100,     # cost of treating flu
  p1=0.05,      # probability of contracting flu under standard care
  p2=0.03       # probability of contracting flu under NI prophylaxis				   
)

#RESULTS

results <- jags(
  data = data,
  # inits = NA,
  parameters.to.save = c("cdrug", "cdoc", "cflu", "p1", "p2",
                         "costtrt", "costctl", "diff.cost"),
  model.file = model_code,
  n.chains = 1,
  n.iter=1
)


results

```


