---
title: 'Chapter 10: Markov models'
author: "Bob"
date: "2025-04-29"
output: html_document
---


NOTE: All the examples for this chapter now compile and run. Those not explicitly noted as matching the results in the text or winBUGS have not been checked against those sources.

So far I've seen three major gotchas for porting the WinBUGS code for this chapter to R/JAGS, listed here with workarounds:

* "Attempt to redefine node" error: try moving assignments to a data block.
* "Error in node (r[1,3,1]/rhat[1,3,1]) Invalid parent values": this is actually a divide-by-zero error in disguise; try adding a really small value to the denominator.
* 3d data needs to be in row-major order: restructure 3D arrays until they match the indexing expected in the JAGS/BUGS code.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')
library('tidyr')      # pivot_longer
library('knitr')      # HTML table display

```


# Example 10.2 revisited (p 203)

This model is hand-typed from the book. The challenge here is that we can no longer pretend that JAGS/BUGS code for data is the same as R code. That is true for lists and matrixes, but not for higher-dimensional arrays, so the data needs to be munged into the row-major order that JAGS expects. 

```{r example_10_2_revisited}

model_code <- "
model{
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      r[tmt, i, 1:5] ~ dmulti(pi[tmt, i, 1:5], n[tmt, i]) # multinomial data
      pi[tmt, i, 1:5] ~ ddirch(prior[tmt, i, 1:5])        # Dirichlet prior
    }
  }

}
" %>% textConnection

# We need to do some munging to create an R data structure with the
# correct dimensions. First we make an R version of the `r` 3d array,
# using the data as it is printed in the text:
rr=structure(
  .Data=c(                    # Seretide
    210,	 60,	0,	1,	 1,	  # Row 1 (STW)
     88,	641,	0,	4,	13,	  # Row 2 (UTW)
      0,		0,	0,	0,	 0,	  # Row 3 (Hex)
      1,		0,	0,	0,	 1,	  # Row 4 (Pex)
    
                              # Fluticasone
     66,   32,	0,	0,	 2,	  # Row 1 (STW)
     42,  752,	0,	5,	20,	  # Row 2 (UTW)
      0,    0,  0,	0,	 0,	  # Row 3 (Hex)
      0,    4,  0,	1,	 0    # Row 4 (Pex)
    ), 
    .Dim=c(5,4,2)             # Matrix dimensions # 2,4,5
)

# The Seretide matrix is t(r[,,1]), and the Fluticasone matrix is t(r[,,2])

# Now we need to shift the values around to the arrangement JAGS expects. 
# Start with a 3d array full of placeholder values:
jr <- structure(
  .Data=rep(-1, 2*4*5),     # -1 can't really happen
  .Dim=c(2,4,5)             # Matrix dimensions
)

# Now copy in the numbers from the R array:
for (tmt in 1:2){
  for (i in 1:4){
    for (j in 1:5){
      jr[tmt, i, j] <- rr[j, i, tmt]
    }
  }
}

## The row totals need to be swapped and flipped (row major order) to match the `r` array.
## Since this is 2d data we can just use a matrix:
jn <- matrix(
    c(272,746,0,2,
      100,819,0,5),
    nrow=2, byrow=TRUE
)

data <- list(
  r=jr,                      # `r` array in row-major order
  n=jn,                      # Row totals
  prior=structure(           # Dirichlet priors all = 1
    .Data=rep(1, 2*4*5), 
		.Dim=c(2,4,5)
	)
)

pi_results <- jags(
  data = data,
  parameters.to.save = "pi",
  model.file = model_code,
  n.chains = 1,
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000
)

```
<style>
  table {
    width: auto;
  }
  td {
    border: 2px solid #999;
    padding: 10px 5px;
  }
  thead {
    background-color: #999;
    border: 2px solid #999;
    color: white;
  }
  tbody {
    color: black;
  }
</style>

```{r pi_smry, results='asis'}
pi_smry <- summary(pi_results)

keeper_cols <- c("mean", "2.5%", "97.5%")

states <- c('STW', 'UTW', 'Hex', 'Pex', 'TF')
treatments = c('Seretide', 'Fluticasone')

seretide_flags <- dimnames(pi_smry)[[1]] %>% grepl("pi[1", ., fixed=TRUE)
fluticasone_flags <- dimnames(pi_smry)[[1]] %>% grepl("pi[2", ., fixed=TRUE)
# fluticasone_flags is not the same as !seretide_flags because of 'deviance'
tt_long <- rbind(
  pi_smry[seretide_flags, keeper_cols],
  pi_smry[fluticasone_flags, keeper_cols]
) %>% 
  as.data.frame %>% 
  mutate(
    value=sprintf("%0.3f<br/>(%0.2f, %0.2f)", mean, `2.5%`, `97.5%`),
  ) %>% select(value)

pat_matches <- row.names(tt_long) %>% regexec("pi\\[(\\d),(\\d),(\\d)\\]", .)
idx_df <- regmatches(row.names(tt_long), pat_matches) %>%
  sapply(function(v) v[2:4] %>% as.numeric) %>% 
  t %>%
  data.frame %>%
  setNames(., nm = c('tmt', 'i', 'j')) %>%
  mutate(
    Treatment = treatments[tmt],
    From = states[i],
    To = states[j]
  ) %>% 
  select(Treatment, From, To)

ttl <- bind_cols(idx_df, tt_long)

ttl %>% 
  pivot_wider(id_cols=c('Treatment', 'From'), names_from=c("To")) %>%
  as.data.frame %>% 
  # xtable %>% print(type = "html")
  knitr::kable(., format='html', escape=FALSE)

```

These numbers correspond well to those in Table 10.4, p 204. I have not included rows showing the probabilities of transitioning from the terminal state 'TF', since that state can only transition to itself.


## Residual Deviance (p 205)

Here we add the code from p205 to compute residual deviance. The data remains the same.

We now encounter a new problem; the values of rhat for state 3 are exactly 0, which gives us a division problem. KLUDGE ALERT: I am addressing this by adding a really small value to rhat so it cannot be 0. I mean REALLY small.


```{r residual_deviance}
model_code <- "
data{
  smidge <- 2.225074e-308 # from `.Machine$double.xmin` in R
  # Yes, it is a kludge, but that is a really small number!
}

model{
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      r[tmt, i, 1:5] ~ dmulti(pi[tmt, i, 1:5], n[tmt, i]) # multinomial data
      pi[tmt, i, 1:5] ~ ddirch(prior[tmt, i, 1:5])        # Dirichlet prior
      ## We could just add the new operations into the existing loops:
      # for (j in 1:5){
      #   rhat[tmt, i, j] <- pi[tmt, i, j] * n[tmt, i] + smidge # Predicted no. given model
      #   dev[tmt, i, j] <- 2 * r[tmt, i, j] * log( r[tmt, i, j]/rhat[tmt, i, j]) # residual deviance for each cell
      # }
      # resdev[tmt, i] <- sum(dev[tmt, i, 1:5])         # residual deviance for each row
    }
  }
  
  # Instead, I use separate loops to emphasize the new code.
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      for (j in 1:5){
        rhat[tmt, i, j] <- pi[tmt, i, j] * n[tmt, i] + smidge # Predicted no. given model
        dev[tmt, i, j] <- 2 * r[tmt, i, j] * log(r[tmt, i, j]/rhat[tmt, i, j]) # residual deviance for each cell
      }
      resdev[tmt, i] <- sum(dev[tmt, i, 1:5])         # residual deviance for each row
    }
  }
  
  resdevtot <- sum(resdev[1:2, 1:4])                  # total residual deviance

}
" %>% textConnection

dev_results <- jags(
  data = data,
  parameters.to.save = c("resdev", "resdevtot"),
  model.file = model_code,
  n.chains = 1,
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000
)

dev_summary <- summary(dev_results)

display_order <- dimnames(dev_summary)[[1]] %>% sort %>% "["(c(2:10, 1))

dev_summary[display_order,]
```

These values are comparable to those in Table 10.5, p205.

# `Example 10.3 revisited: Prophylactic treatments for asthma`

I typed this additional code in from the book (p208), and added it to the code from Example 10.2.

Note that the code for this model is different from that given in the Supplemental Materials (`asthma_1` and `asthma_2` below).

```{r example_10_3_revisited}
model_code <- "
data{
  smidge <- 2.225074e-308 # from `.Machine$double.xmin` in R
  # Yes, it is a kludge, but that is a really small number!
}

model{
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      r[tmt, i, 1:5] ~ dmulti(pi[tmt, i, 1:5], n[tmt, i]) # multinomial data
      pi[tmt, i, 1:5] ~ ddirch(prior[tmt, i, 1:5])        # Dirichlet prior
      ## We could just add the new operations into the existing loops:
      # for (j in 1:5){
      #   rhat[tmt, i, j] <- pi[tmt, i, j] * n[tmt, i] + smidge # Predicted no. given model
      #   dev[tmt, i, j] <- 2 * r[tmt, i, j] * log( r[tmt, i, j]/rhat[tmt, i, j]) # residual deviance for each cell
      # }
      # resdev[tmt, i] <- sum(dev[tmt, i, 1:5])         # residual deviance for each row
    }
  }
  
  # Instead, I use separate loops to emphasize the new code.
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      for (j in 1:5){
        rhat[tmt, i, j] <- pi[tmt, i, j] * n[tmt, i] + smidge # Predicted no. given model
        dev[tmt, i, j] <- 2 * r[tmt, i, j] * log(r[tmt, i, j]/rhat[tmt, i, j]) # residual deviance for each cell
      }
      resdev[tmt, i] <- sum(dev[tmt, i, 1:5])         # residual deviance for each row
    }
  }
  
  resdevtot <- sum(resdev[1:2, 1:4])                  # total residual deviance

  # This last section is Example 10.3 revisited, p 208:
  for (tmt in 1:2){                      # Loop over treatment
    for (i in 1:5){                      # Initialize starting state: 1 in
      s[tmt, i, 1] <- equals(i, 1)       # STW, 0 in all other states
    }
    
    for (i in 1:4){
      for (t in 2:13){                   # Run the model for 12 cycles
        s[tmt, i, t] <- inprod(s[tmt, 1:4, t-1], pi[tmt, 1:4, i]) # Calculate inner product
      }
      E[tmt, i] <- sum(s[tmt, i, 2:13])  # Sum up the time spent in state i
    }
    E[tmt, 5] <- 12 - sum(E[tmt, 1:4])   # Time in TF is 12 weeks
  }                                      # minus time in states 1 to 4.
                                         # Additional expected time in
  for (i in 1:5){
    D[i] <- E[1, i] - E[2, i]
    prob[i] <- step(D[i])
  }
}
" %>% textConnection

results_10_3 <- jags(
  data = data,
  parameters.to.save = c("E", "D", "prob"),
  model.file = model_code,
  n.chains = 1,
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000
)

summary_10_3 <- summary(results_10_3)

summary_10_3
```

Reformat these results for comparison to Table 10.6 (p 209):

```{r reformat_summary_10_3, results='asis'}

summary_10_3 <- summary(results_10_3)

keeper_cols <- c("mean", "2.5%", "97.5%")

states <- c('STW', 'UTW', 'Hex', 'Pex', 'TF')
treatments = c('Seretide', 'Fluticasone')

# Expected number of weeks in state
E_seretide_flags <- dimnames(summary_10_3)[[1]] %>% grepl("E[1", ., fixed=TRUE)
E_fluticasone_flags <- dimnames(summary_10_3)[[1]] %>% grepl("E[2", ., fixed=TRUE)
expected_weeks <- rbind(
    summary_10_3[E_seretide_flags, keeper_cols],
    summary_10_3[E_fluticasone_flags, keeper_cols]
  ) %>% 
  as.data.frame %>% 
  mutate(
    treatment=c(rep('Seretide', sum(E_seretide_flags)), rep('Fluticasone', sum(E_fluticasone_flags))),
    state=c(states, states),
    value=sprintf("%0.2f", mean),
  ) %>% 
  select(treatment, state, value) %>%
  pivot_wider(id_cols='treatment', names_from='state')

expected_weeks %>%
  as.data.frame %>% 
  knitr::kable(., format='html', escape=FALSE, caption="Expected number of weeks in state")

# Additional expected time in state for Seretide compared to Fluticasone
D_flags <- dimnames(summary_10_3)[[1]] %>% grepl("D[", ., fixed=TRUE)
summary_10_3[D_flags, c("mean", "2.5%", "97.5%")] %>% 
  apply(1, function(v) sprintf("%0.2f (%0.1f, %0.1f)", v[1], v[2], v[3])) %>% 
  setNames(., nm=states) %>% as.data.frame %>% t %>%
  knitr::kable(., format='html', escape=FALSE, caption="Additional expected time in state for Seretide compared to Fluticasone")

# Probability that expected time in state is greater for Seretide than Fluticasone
prob_flags <- dimnames(summary_10_3)[[1]] %>% grepl("prob[", ., fixed=TRUE)
summary_10_3[prob_flags, "mean"] %>% 
  setNames(., nm=states) %>% sapply(function(v) sprintf("%0.3f", v[1])) %>% as.data.frame %>% t %>%
  knitr::kable(., format='html', escape=FALSE, caption="Probability that expected time in state is greater for Seretide than Fluticasone")

```

These results are comparable to those in Table 10.6 (p 209).


# `asthma_1`

Estimating Transition Probabilities

```{r asthma_1}

# Model

model_code <- "
data{
  smidge <- 2.225074e-308
}
model{
	for (tmt in 1:2){						# Treatments tmt=1 (Seretide), tmt=2 (Fluticasone)
		for (i in 1:4){						# There are 4 non-absorbing health states
			r[tmt, i, 1:5] ~ dmulti(pi[tmt,i,1:5], n[tmt,i])		# Multinomial data
			pi[tmt, i, 1:5] ~ ddirch(prior[tmt,i,1:5])		# Dirichlet prior for probs.
			for (j in 1:5){
				rhat[tmt,i,j] <- pi[tmt,i,j] * n[tmt,i] + smidge
				dev[tmt,i,j] <- 2*r[tmt,i,j] * log(r[tmt,i,j]/rhat[tmt,i,j])
			}
			resdev[tmt,i] <- sum(dev[tmt,i,1:5])
		}
	}

  resdevtot <- sum(resdev[1:2,1:4])
}

" %>% textConnection

## Modify data from earlier model
data$n <- matrix(c(rowSums(data$r[1,,]), rowSums(data$r[2,,])), nrow=2, byrow=TRUE)
data$prior=structure(
    .Data=rep(1, 2*4*5),
		.Dim=c(2,4,5))

# Data: same as in model above

# Pretend the WinBugs data structure is R:
rpi <- structure(
        .Data=c(.6,.1,.1,.1,.1,
                .1,.6,.1,.1,.1,
                .1,.1,.6,.1,.1,
                .1,.1,.1,.6,.1,
  
      					.6,.1,.1,.1,.1,
                .1,.6,.1,.1,.1,
                .1,.1,.6,.1,.1,
                .1,.1,.1,.6,.1)
        ,
      	.Dim=c(5,4,2) # reverse the dimensions
      )

# Swap to row-major order for JAGS. Start with a 3d array full of placeholder values:
jpi <- structure(
  .Data=rep(-1, 2*4*5),     # -1 can't really happen
  .Dim=c(2,4,5)             # Matrix dimensions
)

# Now copy in the numbers from the R array:
for (tmt in 1:2){
  for (i in 1:4){
    for (j in 1:5){
      jpi[tmt, i, j] <- rpi[j, i, tmt]
    }
  }
}

# Starting/initial values: same as random effects model without sd
initial_values <- list(
  # Inits
  list( pi=jpi )
  ,
  
  # Alternative Inits
  list(
    pi=structure(
      .Data=rep(0.2, 2*4*5),
  	  .Dim=c(2,4,5)
    )
  )
  
)


parameters_to_save <- c("r", "pi", "rhat") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```


# `asthma_2`

Propagating Uncertainty into a Decision Model


```{r asthma_2}


# Model



model_code <- "
model{
# Data analysis
	for (tmt in 1:2){										# Treatments tmt=1 (Seretide), tmt=2 (Fluticasone)
		for (i in 1:4){									  # There are 4 non-absorbing health states
			r[tmt,i,1:5] ~ dmulti(pi[tmt,i,1:5],n[tmt,i])		# Multinomial data
			pi[tmt,i,1:5] ~ ddirch(prior[tmt,i,1:5])		    # Dirichlet prior for probs.
		}
	}
# Calculating summaries from a decision model
	for (tmt in 1:2){ 
		for (i in 1:5){ s[tmt,i,1]<- equals(i,1) }  	# Initialise starting state: 1 in STW, 0 in all other states
		for (i in 1:4){  
			for (t in 2:13){
				s[tmt,i,t] <- inprod(s[tmt,1:4,t-1], pi[tmt,1:4,i]) 	# Run the model for 12 cycles. 
																			# s[tmt,i,t] = no. in state i at time t under treatment tmt
				}
			E[tmt,i] <- sum(s[tmt,i,2:13])							# Sum up time spent in state i
			}
		E[tmt,5] <- 12 - sum(E[tmt,1:4])							# Time in TF = 12 minus time in other states.
	}

	for (i in 1:5){
		D[i] <- E[1,i] - E[2,i]									# Additional time in state i under Seretide rather than FT
		prob[i] <- step(D[i])										# Indicates whether Seretide gives longer time in state i
	}
		
}

" %>% textConnection

data$prior <- structure( .Data=rep(1, 2*4*5), .Dim=c(2,4,5))
# data <- list(
# r=structure(.Data=c(210,60,0,1,1,	88,641,0,4,13,	0,0,0,0,0,	1,0,0,0,1,	
# 							66,32,0,0,2,	42,752,0,5,20,	0,0,0,0,0,	0,4,0,1,0), 
# 				.Dim=c(2,4,5)),
# n=structure(.Data=c(272,746,0,2,	100,819,0,5), .Dim=c(2,4)),
# prior=structure( .Data=rep(1, 2*4*5), .Dim=c(2,4,5))
# )


initial_values <- list(
  #Inits
  list(pi=structure(.Data=c(.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1,	
  								.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1), 
  				.Dim=c(2,4,5))
  ),
  
  #Alternative Inits
  list(pi=structure(.Data=c(.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,
  								.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2),
  				.Dim=c(2,4,5))
  )

  
)

parameters_to_save <- c("r", "pi", "s", "E", 'D', 'prob') # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

summary(results)

```

# Estimating transition parameters from a synthesis of several studies

## Modeling study effects

Random effects model for $\lambda_{1,s}$

This is the example from p 214. 

<font color="#F00">
<h2>Why won't this give me values for lambda1?</h2>
</font>

```{r lambda_1_s}
model_code <- "
model {
  for (s in 1:5) {
    loglam1[s] ~ dnorm(L, prec)
    
    log(lambda1[s]) <- loglam1[s]
    gamma12[s] <- lambda1[s] * (1 - rho)  # !!! The book has [i] instead of [s] !!!
    gamma13[s] <- lambda1[s] * rho
  }
  
  gamma23 ~ dexp(0.001)
  L ~ dnorm(0, 0.00001)
  tau ~ dunif(0,1)
  prec <- 1/(tau*tau)
  rho ~ dbeta(1,1)
  
  foohat ~ dnorm(foo, 1) # Avoid message about unused variable 'foo'
}

" %>% textConnection

parameters_to_save <- c(
  "L", "loglam1", "lambda1", # "gamma12", "gamma13",
  "gamma23", 'foohat'
  )

results <- jags(
  data = list(foo=99),  # this wrapper function really wants some data
  # inits = initial_values,
  parameters.to.save = parameters_to_save,
  model.file = model_code,
  n.chains = 1,
  n.adapt = 100,
  n.iter = 20000,
  n.burnin = 10000)

summary(results)

```

## Synthesis of studies reporting aggregate data

This is the code from pp 216-217.

## Incorportating studies that provide event history data

Add the code from p 219.

## Reporting results from a Random Effects Model

Add the code from p 220.

## Incorporating Treatment Effects

Add the code from p 223.


# `evsyn_markov_book_1`
Synthesis of Studies Reporting Aggregate and Event History Data

I had to make two kinds of changes to get this to run:

* move the value assignment for n[i] to the data block
* rearrage the 3d data in r.agg, then make sure the sums in n.agg match the values in r.agg

<font color="#F00">
TO DO: Check these answers against WinBUGS.
</font>


```{r evsyn_markov_book_1}

model_code <- "

data {
  for (i in 1:2) {
      n[i] <- m12[i] + m13[i]
  }
}

model {

  # MODEL

  # Random effects model for logs of study-specific rates lamda1 of leaving state 1
  for (s in 1:15) {
    loglam1[s] ~ dnorm(L,prec)        # Random effects model for lambda1
    log(lambda1[s]) <- loglam1[s]
    gamma12[s] <- lambda1[s]*(1-rho)  # Study specific transition rates 1 to 2
    gamma13[s] <- lambda1[s]*rho      # Study specific transition rates 1 to 3
  }   

  # PRIORS
  
  gamma23 ~ dexp(.001)              # vague prior for 'fixed' transition rate 2 to 3
  L ~ dnorm(0,.00001)               # vague prior for population mean loglamda1
  tau ~ dunif(0,1)                  # vague prior for sd for population loglamda1
  prec <- 1/(tau*tau)  
  rho ~ dbeta(1,1)                  # vague prior for proportion going to state 2 from 1
  
  
  # LIKELIHOOD 

  # 2 studies reported aggregated total no. of transitions 
	for (s in 1:2){
		for (i in 1:2){
			r.agg[s,i,1:3] ~ dmulti(pi.agg[s,i,1:3], n.agg[s,i])
		}
	}
	
	for (obs in 1:6){
	 	pi.agg[obs,1,1] <- exp(-(gamma12[study[obs]] + gamma13[study[obs]])*T[obs])
	 	pi.agg[obs,1,2] <- gamma12[study[obs]]*(exp(-gamma23 * T[obs]) - exp(- (gamma12[study[obs]]+gamma13[study[obs]])*T[obs] ) ) /((gamma12[study[obs]]+gamma13[study[obs]]) - gamma23)
	 	pi.agg[obs,1,3] <- 1 - pi.agg[obs,1,1] - pi.agg[obs,1,2]
	
	 	pi.agg[obs,2,1] <- 0
	 	pi.agg[obs,2,2] <- exp(-gamma23)
	 	pi.agg[obs,2,3] <- 1 - pi.agg[obs,2,2]
	}

  # 3 studies (4 data points) reported numbers number starting in state 1 and entering state 3   
  # during time T. Study 4 consists of 2 groups of individuals with different follow-up times, T
  for (i in 1:4) {
    r.d[i] ~ dbin(pi.agg[(i+2),1,3], n.d[i]) 
  }

  # 8 studies reporting event history data informing individual transitions
  for (i in 1:2) {
    r12[i] ~ dpois(mu[i+5])
    log(mu[i+5]) <- log(E12[i]) + log(gamma12[i+5])
  }
  for (i in 1:2) {
    r13[i] ~ dpois(mu[i+7])
    log(mu[i+7]) <- log(E13[i]) + log(gamma13[i+7])
  }
  for (i in 1:4) {
    r23[i] ~ dpois(mu[i+9])
    log(mu[i+9]) <- log(E23[i]) + log(gamma23)
  }

  # 2 studies reporting event history data on first tranistion from state 1
  for (i in 1:2) {
    # n[i] <- m12[i] + m13[i] #!!! to data block
    n[i] ~ dpois(nu[i])                      # likelihood for Poisson rate
    log(nu[i]) <- log(E[i]) + loglam1[i+13]  
    m13[i] ~ dbin(rho,n[i])                  # likelihood for binomial proportion going to state 3
  }

  # Three different ways to summarise random effects model for lambda1, and form transition #rates. 
  
  # 1. Use population mean, L
  g12[1] <- (1-rho) * exp(L)
  g13[1] <- rho * exp(L)
  
  # 2. Use predictive distribution for loglam1
  loglam1.pred ~ dnorm(L,prec)
  log(lambda1.pred) <- loglam1.pred
  g12[2] <- (1-rho) * lambda1.pred
  g13[2] <- rho * lambda1.pred
  
  # 3. Calibrate to study 15
  g12[3] <- (1-rho) * lambda1[15]
  g13[3] <- rho * lambda1[15]

  # Transition probabilities over observation period T=1, for each RE summary
  for (j in 1:3){
  	P11[j] <- exp(-(g12[j] + g13[j]))
  	P12[j] <- g12[j]*(exp(-gamma23) - exp(-(g12[j] + g13[j])))/(g12[j] + g13[j] - gamma23)
  	P13[j] <- (1.0 - P11[j] - P12[j])
  	P22[j] <- exp(-gamma23)
  	P23[j] <- 1 - P22[j]
  }

}

" %>% textConnection

# DATA LIST
data <- list(                                                                                           

  #STUDIES 1-2
  #2 studies reporting aggregate data for all states, for given observation cycle, T
  
  ## !!! This 3D array needs to be rearranged; it will be added below
	# r.agg=structure(.Data=c(87,	4,	9,	
	# 					    					0,	3,	1,
	# 					
	# 											210,	8,	15,
	# 											0,		5,		3), .Dim=c(2,2,3)), 

	# n.agg=structure(.Data=c(100,	4,
	# 											233,	8), .Dim=c(2,2)), 

  # STUDIES 3-5
  # 3 studies reporting no. transitions to death for given observation cycle, T
  r.d=c(38,15,11,60),      n.d=c(181,177,103,335) , 

  #Observation time & study number for 6 aggregate data observations
  T=c(1,1, 2,1.5,2,5), study=c(1,2,3,4,4,5),

  # STUDIES 6-13
  # 8 studies with event history data on individual transitions
  r12=c(8,20),             E12=c(120,620),
  r13=c( 12,44),           E13=c(140,677),  
  r23=c(9,12,5,6),         E23=c(34,35,15,25),

  # STUDIES 14-15
  # 2 studies with event history data on first transition from state 1
  m12=c(18,30),
	m13=c(40,75),
	E=c(380,1169) #!!! , trailing comma was OK in WinBUGS     

)

studies <- list(
  matrix(c(87,	4,	9,	
	  				0,	3,	1), nrow=2, byrow=TRUE),
  matrix(c(210,	8,	15,	
	  				 0,	5,	 3), nrow=2, byrow=TRUE)
)
data$r.agg <- structure( .Data=rep(-1, 2*2*3), .Dim=c(2,2,3))
data$n.agg=structure(.Data=rep(-1, 2*2), .Dim=c(2,2))
for (s in 1:2){
  for (r in 1:2){
    for (c in 1:3){
      data$r.agg[s, r, c] <- studies[[s]][r,c]
    }
    data$n.agg[s, r] <- sum(data$r.agg[s, r, 1:3])
  }
}

# INITIAL VALUES    

initial_values <- list(
  list(tau=.2,   rho=.4,    gamma23=.02,    L=-3,
       loglam1=c(.02,.02,.02,.02,.02,   .02,.02,.02,.02, .02,   .02,.02,.02,.02,.02),loglam1.pred=.02),
  
  list(tau=.1,   rho=.5,    gamma23=.08,    L=-2,
       loglam1=c(.01,.03,.01,.03,.01,   .01,.03,.01,.01, .03,   .03,.01,.02,.01,.03),loglam1.pred=.03)

)

parameters_to_save <- c("P11", "P12", "P13", "P22", "P23") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

summary(results)

```


# `evsyn_markov_book_2`
Including Studies with Treatment Comparisons

When I tried to run the WinBUGS code directly, I got this error:
<pre>
RUNTIME ERROR:
Compilation error on line 70.
Attempt to redefine node n[1]
</pre>

I fixed that by moving one line of code to a data block. Then I got:
<pre>
  Error in node r.aggt[1,1,1:3]
Node inconsistent with parents
</pre>

This was fixed by rearranging the 3d data structures.


```{r evsyn_markov_book_2}
model_code <- "
data{
  smidge <- 2.225074e-308
  for (i in 1:2) {
    n[i] <- m12[i] + m13[i] + smidge # !!! to data block
  }
}

model {

#MODEL

# Random effects model for logs of study-specific rates lamda1 of leaving state 1
for (s in 1:15) { 
  loglam1[s] ~ dnorm(L,prec) 	   #  Random effects model for lambda1
	log(lambda1[s]) <- loglam1[s]
	gamma12[s]<-lambda1[s]*(1-rho) # Study specific transition rates 1 to 2
	gamma13[s]<-lambda1[s]*rho	   # Study specific transition rates 1 to 3
}   

# PRIORS

gamma23 ~ dexp(.001)             # vague prior for 'fixed' transition rate 2 to 3
L ~ dnorm(0,.00001)              # vague prior for population mean loglamda1
tau ~ dunif(0,1)                 # vague prior for sd for population loglamda1
prec <- 1/(tau*tau)  
rho ~ dbeta(1,1)                 # vague prior for proportion going to state 2 from 1

logh~dnorm(0,.00001)						 # Normal prior on log-hazard ratio
log(h)<- logh

# LIKELIHOOD 

# 2 studies reported aggregated total no. of transitions 
	for (s in 1:2){
		for (i in 1:2){
			r.agg[s,i,1:3] ~ dmulti(pi.agg[s,i,1:3],n.agg[s,i])
			r.aggt[s,i,1:3] ~ dmulti(pi.aggt[s,i,1:3],n.aggt[s,i])
		}
	}

	for (s in 1:2){
		pi.aggt[s,1,1] <- exp(-(gamma12[s]*h+gamma13[s])*T[s])
		pi.aggt[s,1,2] <- gamma12[s]*h*(exp(-gamma23 * T[s])  
		- exp(-(gamma12[s]*h+gamma13[s])*T[s] ) ) /(gamma12[s]*h+gamma13[s] - gamma23 )
		pi.aggt[s,1,3]<- 1 - pi.aggt[s,1,1] - pi.aggt[s,1,2]
	
		pi.aggt[s,2,1]<- 0
		pi.aggt[s,2,2]<- exp(-gamma23)
		pi.aggt[s,2,3]<- 1 - pi.aggt[s,2,2]
	}

	for (obs in 1:6){
		pi.agg[obs,1,1]<- exp(-(gamma12[study[obs]]+gamma13[study[obs]])*T[obs])
		pi.agg[obs,1,2]<- gamma12[study[obs]]*(exp(-gamma23 * T[obs])  - exp(- (gamma12[study[obs]]+gamma13[study[obs]])*T[obs] ) ) /((gamma12[study[obs]]+gamma13[study[obs]])- gamma23)
		pi.agg[obs,1,3]<- 1 - pi.agg[obs,1,1] - pi.agg[obs,1,2]
	
		pi.agg[obs,2,1]<- 0
		pi.agg[obs,2,2]<- exp(-gamma23)
		pi.agg[obs,2,3]<- 1 - pi.agg[obs,2,2]
	}

  # 3 studies (4 data points) reported numbers number starting in state 1 and entering state 3 #during time T. Study 4 consists of 2 groups of individuals with different follow-up times, T
  for (i in 1 : 4) {
    r.d[i] ~ dbin(pi.agg[(i+2),1,3],n.d[i]) 
  }
  
  # 8 studies reporting event history data informing individual transitions
  for (i in 1:2) {r12[i] ~dpois(mu[i+5])
                        log(mu[i+5]) <- log(E12[i])  +  log(gamma12[i+5])  }
  for (i in 1:2) {r13[i] ~dpois(mu[i+7])
                        log(mu[i+7]) <- log(E13[i])  +  log(gamma13[i+7])   }
  for (i in 1:4) {r23[i] ~dpois(mu[i+9])
                        log(mu[i+9]) <- log(E23[i])  +  log(gamma23)     }
  
  # 2 studies reporting event history data on first tranistion from state 1
  for (i in 1:2) {
    # n[i] <- m12[i] + m13[i] # !!! to data block
    n[i] ~ dpois(nu[i])                       # likelihood for Poisson rate
    log(nu[i]) <- log(E[i]) +  loglam1[i+13]  
    m13[i] ~ dbin(rho,n[i])                   # likelihood for binomial proportion going to state 3
  }
  
  # Three different ways to summarise random effects model for lambda1, and form transition #rates. 
  
  # 1. Use population mean, L
  g12[1] <- (1-rho) * exp(L)
  g13[1] <-   rho * exp(L)
  
  # 2. Use predictive distribution for loglam1
  loglam1.pred~dnorm(L,prec)
  log(lambda1.pred)<-loglam1.pred
  g12[2] <- (1-rho) * lambda1.pred
  g13[2] <- rho * lambda1.pred
  
  # 3. Calibrate to study 15
  g12[3] <- (1-rho) * lambda1[15]
  g13[3] <- rho * lambda1[15]
  
  # Transition probabilities over observation period T=1, for each RE summary
  for (j in 1:3){
  	P11[j] <- exp(-(g12[j]+g13[j]))
  	P12[j] <- g12[j]*(exp(-gamma23)-exp(-(g12[j]+g13[j])))/(g12[j]+g13[j]-gamma23)
  	P13[j] <- (1-P11[j]-P12[j])
  	P22[j] <- exp(-gamma23)
  	P23[j] <- 1-P22[j]
  }

}

" %>% textConnection

# DATA LIST
data <- list(                                                                                         

# !!! The 3d structures r.agg and r.aggt need to be rearranged, and their 2d row sums need to match.
# !!! This is done in code below.
# # STUDIES 1-2
# # 2 studies reporting aggregate data for all states, for given observation cycle, T
# 	r.agg=structure(.Data=c(87,	4,	9,	
# 						    					0,	3,	1,
# 						
# 												210,	8,	15,
# 												0,		5,		3), .Dim=c(2,2,3)), 
# 	n.agg=structure(.Data=c(100,	4,
# 												233,	8), .Dim=c(2,2)), 
# 
# # Studies 1-2 also report a treatment arm
# 	r.aggt=structure(.Data=c(140,	2,	14,	
# 						    					0,	2,	0,
# 						
# 												264,	5,	27,
# 												0,		3,		2), .Dim=c(2,2,3)), 
# 	n.aggt=structure(.Data=c(156,	2,
# 												296,	5), .Dim=c(2,2)), 

# STUDIES 3-5
# 3 studies reporting no. transitions to death for given observation cycle, T
       r.d=c(38,15,11,60),      n.d=c(181,177,103,335) , 

# Observation time & study number for 6 aggregate data observations
T=c(1,1, 2,1.5,2,5), study=c(1,2,3,4,4,5),

# STUDIES 6-13
#  8 studies with event history data on individual transitions
   r12=c(8,20),               E12=c(120,620),
   r13=c( 12,44),            E13=c(140,677),  
   r23=c(9,12,5,6),         E23=c(34,35,15,25),

#STUDIES 14-15
# 2 studies with event history data on first transition from state 1
     m12=c(18,30),     m13=c(40,75),         E=c(380,1169)     

)

studies <- list(
  matrix(c(87,	4,	9,	
	  				0,	3,	1), nrow=2, byrow=TRUE),
  matrix(c(210,	8,	15,	
	  				 0,	5,	 3), nrow=2, byrow=TRUE)
)
treatment_arm <- list(
  matrix( c(140,	2,	14,	
  						0,	2,	0), nrow=2, byrow=TRUE),
  matrix( c(264,	5,	27,
  						0,	3,	 2), nrow=2, byrow=TRUE)
)
data$r.agg <- data$r.aggt <- structure( .Data=rep(-1, 2*2*3), .Dim=c(2,2,3))
data$n.agg <- data$n.aggt <- structure(.Data=rep(-1, 2*2), .Dim=c(2,2))
for (s in 1:2){
  for (r in 1:2){
    for (c in 1:3){
      data$r.agg[s, r, c] <- studies[[s]][r,c]
      data$r.aggt[s, r, c] <- treatment_arm[[s]][r,c]
    }
    data$n.agg[s, r] <- sum(data$r.agg[s, r, 1:3])
    data$n.aggt[s, r] <- sum(data$r.aggt[s, r, 1:3])
  }
}


# INITIAL VALUES
initial_values <- list(
  list(tau=.2,   rho=.4,    gamma23=.02,    L=-3,
       loglam1=c(.02,.02,.02,.02,.02,   .02,.02,.02,.02, .02,   .02,.02,.02,.02,.02),loglam1.pred=.02,logh=0),
  
  list(tau=.1,   rho=.5,    gamma23=.08,    L=-2,
       loglam1=c(.01,.03,.01,.03,.01,   .01,.03,.01,.01, .03,   .03,.01,.02,.01,.03),loglam1.pred=.03,logh=-1)
)

parameters_to_save <- c("P11", "P12", "P13", "P22", "P23") # What else?


results <- jags(data = data,
                inits = initial_values,
                parameters.to.save = parameters_to_save,
                model.file = model_code,
                n.chains = length(initial_values),
                n.adapt = 100,
                n.iter = 50000,
                n.burnin = 20000)

summary(results)

```

# To Do: 

* Enter the code on the following pages: 216-217, 219, 220, 223
* Compare to the results in the book
* Compare to the programs from the Supplemental Materials
* Where do the models from the Supplemental Materials (`evsyn_markov_book_1` and `evsyn_markov_book_2`) fit in? Note that they use for "s in 1:15")