---
title: 'Chapter 11: Generalized evidence synthesis'
author: "Bob"
date: "2025-03-02"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')
library('knitr')      # kable
```



# Models discussed in the text

## `HRT_power_prior`

This is essentially the model from Box 11.1 (pp231-232), except the code in the text does ot compute `prob.OR1[k]`.

HRT - Power Prior Model

Adjusted for duration - centred at 5 years

Prior based on distribution of pooled effect (Lancet 1997)

  OR 1.31  & 95%CI: 1.16 to 1.46
  => logOR=0.27 & 95% CI: 0.144 to 0.382
  => SD(logOR) = 0.061 => precision = 271.3


The original WinBUGS code gave an `Attempt to redefine node` errors in JAGS, which was addressed by moving the dataset copies to a data block. Here is a copy of the WinBUGS version:
```
model  {
  # Create multiple datasets
  for (i in 1:19){
    for(k in 1:12){   
      rc[i,k] <- rc.dat[i]
      rt[i,k] <- rt.dat[i]
      nc[i,k] <- nc.dat[i]
      nt[i,k] <- nt.dat[i]
    }
  }
  
  # Estimate meta-analysis model for each value of alpha
  for (k in 1: 12) {
    for( i in 1 : 19 ) {
      rc[i,k] ~ dbin(pc[i,k], nc[i,k])
      rt[i,k] ~ dbin(pt[i,k], nt[i,k])
      logit(pc[i,k]) <- mu[i,k]
      logit(pt[i,k]) <- mu[i,k] + delta[i,k]+ beta*(duration[i]-5)
      mu[i,k] ~ dnorm(0.0,1.0E-6)
      delta[i,k] ~ dnorm(d[k], prec[k])
    }					
    d[k] ~ dnorm(0.270, prec.d[k])  # Use mean of meta-analysis of obs data
    OR.5years[k]<- exp(d[k])
    prob.OR1[k] <- step(d[k])  # probability that OR>1, i.e. increase in risk of breast ca  
    prec[k] <- 1/tau.sq[k]
    tau.sq[k] <- tau[k]*tau[k]
    tau[k] ~ dunif(0,5) 
  }

  # Calculate precision  of prior (from Lancet meta-analysis of observational studies) 
  # downweighted using alpha

  for (k in 1:12) {
    prec.d[k] <- alpha[k]*271.3
  }
  beta ~ dnorm(0.0,1.0E-6)
}

```

```{r HRT_power_prior, eval=TRUE}

# Model

model_code <- "

data {
  # Create multiple datasets
  for (i in 1:19){
    for(k in 1:12){   
      rc[i,k] <- rc.dat[i]
      rt[i,k] <- rt.dat[i]
      nc[i,k] <- nc.dat[i]
      nt[i,k] <- nt.dat[i]
    }
  }
}

model  {
  
  # Estimate meta-analysis model for each value of alpha
  for (k in 1: 12) {
    for( i in 1 : 19 ) {
      rc[i,k] ~ dbin(pc[i,k], nc[i,k])
      rt[i,k] ~ dbin(pt[i,k], nt[i,k])
      logit(pc[i,k]) <- mu[i,k]
      logit(pt[i,k]) <- mu[i,k] + delta[i,k]+ beta*(duration[i]-5)
      mu[i,k] ~ dnorm(0.0,1.0E-6)
      delta[i,k] ~ dnorm(d[k], prec[k])
    }					
    d[k] ~ dnorm(0.270, prec.d[k])  # Use mean of meta-analysis of obs data
    OR.5years[k]<- exp(d[k])
    prob.OR1[k] <- step(d[k])  # probability that OR>1, i.e. increase in risk of breast ca  
    prec[k] <- 1/tau.sq[k]
    tau.sq[k] <- tau[k]*tau[k]
    tau[k] ~ dunif(0,5) 
  }

  # Calculate precision  of prior (from Lancet meta-analysis of observational studies) 
  # downweighted using alpha

  for (k in 1:12) {
    prec.d[k] <- alpha[k]*271.3
  }
  beta ~ dnorm(0.0,1.0E-6)
}
" %>% textConnection


# Data

data <- list(
  rt.dat = c(0,2,1,0,1,0,0,2,7,6,1,32,2,1,0,1,0,2,5),
  nt.dat = c(84,56,116,62,39,31,30,100,701,1128,37,1380,232,150,215,204,243,502,337),
  rc.dat = c(4,1,1,1,1,1,1,0,1,0,0,25,1,1,1,0,4,5,5),
  nc.dat = c(84,259,40,20,36,70,30,51,174,137,37,1383,232,50,106,105,246,504,327),
  duration=c(10,2,1,2,1,2.92,1,2,3,2,2,4.08,5,2,1,3.17,3,4.58,2.83),
  alpha=c(0.000001,0.001,0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9,1.0)
)

# Starting/initial values: same as random effects model without sd
initial_values <- list(
  list(
    mu=structure( 
      .Data=rep(-5, 19 * 12), 
      .Dim=c(19,12)
    ),
    delta=structure( 
      .Data=rep(-0.54, 19 * 12), 
      .Dim=c(19,12)
    ),
    d = rep(-0.51, 12),
    beta=-0.19,
    tau=rep(1, 12)
  )
)

parameters_to_save <- c("OR.5years", "prob.OR1")


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

hrt_pp_results_summary <- summary(results)

hrt_pp_results_summary %>% kable
```

### WinBUGS results for HRT power prior model

```{r hrt_power_prior_winbugs_results}

hrt_ppwb_results <- "
node	 mean	 sd	 MC error	2.5%	median	97.5%	start	sample
OR.5years[1]	0.8338	0.4129	0.01195	0.2669	0.7939	1.692	20001	10000
OR.5years[2]	0.8239	0.3979	0.01004	0.2994	0.7648	1.695	20001	10000
OR.5years[3]	1.181	0.207	0.006385	0.8387	1.162	1.656	20001	10000
OR.5years[4]	1.235	0.1614	0.00419	0.947	1.225	1.579	20001	10000
OR.5years[5]	1.257	0.1352	0.002739	1.016	1.247	1.548	20001	10000
OR.5years[6]	1.27	0.1193	0.002548	1.06	1.263	1.529	20001	10000
OR.5years[7]	1.276	0.105	0.002092	1.085	1.271	1.502	20001	10000
OR.5years[8]	1.279	0.09981	0.002264	1.096	1.274	1.488	20001	10000
OR.5years[9]	1.285	0.09417	0.002484	1.115	1.282	1.477	20001	10000
OR.5years[10]	1.29	0.08679	0.001553	1.128	1.287	1.466	20001	10000
OR.5years[11]	1.29	0.0809	0.001679	1.14	1.287	1.458	20001	10000
OR.5years[12]	1.29	0.07937	0.002053	1.139	1.289	1.449	20001	10000
prob.OR1[1]	0.2537	0.4351	0.01604	0.0	0.0	1.0	20001	10000
prob.OR1[2]	0.2292	0.4203	0.01169	0.0	0.0	1.0	20001	10000
prob.OR1[3]	0.8087	0.3933	0.009994	0.0	1.0	1.0	20001	10000
prob.OR1[4]	0.9383	0.2406	0.006712	0.0	1.0	1.0	20001	10000
prob.OR1[5]	0.9826	0.1308	0.001813	1.0	1.0	1.0	20001	10000
prob.OR1[6]	0.9942	0.07594	7.808E-4	1.0	1.0	1.0	20001	10000
prob.OR1[7]	0.9988	0.03462	3.562E-4	1.0	1.0	1.0	20001	10000
prob.OR1[8]	0.9992	0.02827	2.727E-4	1.0	1.0	1.0	20001	10000
prob.OR1[9]	0.9999	0.009999	1.0E-4	1.0	1.0	1.0	20001	10000
prob.OR1[10]	0.9999	0.009999	1.0E-4	1.0	1.0	1.0	20001	10000
prob.OR1[11]	1.0	0.0	1.0E-12	1.0	1.0	1.0	20001	10000
prob.OR1[12]	1.0	0.0	1.0E-12	1.0	1.0	1.0	20001	10000
" %>% read.table(text=., sep="\t", header=TRUE, check.names=FALSE)

# rename = c(X2.5.="2.5%", `X97.5.`= "97.5%")
# for (old_name in names(rename)){
#   new_name <- rename[old_name]
#   names(ppwb_results)[names(ppwb_results)==old_name] <- new_name
# }
hrt_ppwb_results %>% kable
```


## `HRT_bias_allowance`

This is the code from Box 11.2 (pp243-235), with some of the lines in a different order. The code in the text also does not compute `prob.OR1`.

HRT Bias Allowance Model

Adjusted for duration - centred at 5 years

Summary of observational evidence (Lancet 1997)

  OR 1.31  & 95%CI: 1.16 to 1.46
  => logOR=0.27 & 95% CI: 0.144 to 0.382
  => SD(logOR) = 0.061 => var = 0.0037 + var.bias

Var.bias derived from 19 meta-analyses of empirical estimates of 
studies comparing RCTs and non-randomised studies


```{r HRT_bias_allowance}

# Model

model_code <- "model {
  for(j in 1 : Nstud) {
    rc[j] ~ dbin(pc[j], nc[j])
    rt[j] ~ dbin(pt[j], nt[j])
    logit(pc[j]) <- mu[j]
    logit(pt[j]) <- mu[j] + delta[j] + beta*(duration[j]-5)
    mu[j] ~ dnorm(0.0,1.0E-5)
    delta[j] ~ dnorm(d, prec)			
  }

  # Meta-meta-analysis of 19 meta-analyses comparing RCTs and 
  # NRCTs (Ioannidis et al 2001)
  
  for( j in 1 : Nma) {
    logRRR [j] ~ dnorm(d.bias, prec.bias)
  }

  # Priors
  
  prec.bias <- 1/tau.sq.bias
  tau.sq.bias <- tau.bias*tau.bias
  
  d ~ dnorm(0.27, prec.d)
  prec.d <- 1/tau.sq.d
  tau.sq.d <- 0.0037 + tau.sq.bias
  
  prec <- 1/tau.sq
  tau.sq <- tau*tau
  
  tau ~ dunif(0,5)
  tau.bias ~ dunif(0,5)
  d.bias ~ dnorm(0.0,1.0E-5)
  beta ~ dnorm(0.0,1.0E-5)
  OR<- exp(d)
  prob.OR1 <- step(d)
}
	
" %>% textConnection


# Data

data <- list(
  rt = c(0,2,1,0,1,0,0,2,7,6,1,32,2,1,0,1,0,2,5),
  nt = c(84,56,116,62,39,31,30,100,701,1128,37,1380,232,150,215,204,243,502,337),
  rc = c(4,1,1,1,1,1,1,0,1,0,0,25,1,1,1,0,4,5,5),
  nc = c(84,259,40,20,36,70,30,51,174,137,37,1383,232,50,106,105,246,504,327),
  duration=c(10,2,1,2,1,2.92,1,2,3,2,2,4.08,5,2,1,3.17,3,4.58,2.83), 
  Nstud = 19,
  Nma=19,
  logRRR = c(-1.406497068, -0.182721637, 0.299363577, 0.087094707,
    0.805135683, 0.017839918, -0.167235919, 0.133656385, -0.038740828, 0.969262617, 0.083421608,
    -0.432322562, 0, 0.347129531, 0.038258712, -0.066139803, 0.206200831, 0.1889661, -0.534435489)
)

# Starting/initial values: same as random effects model without sd
initial_values <- list(
  list(d = 0, 
       tau=1, 
       mu = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  		 delta = c(0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0, 0),
  		 d.bias=0,  
  		 tau.bias=1, 
  		 beta=0
  )
)

parameters_to_save <- c("OR", "beta", "d", "prob.OR1", "tau", "tau.bias")


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)


hrt_ba_results_summary <- summary(results)

hrt_ba_results_summary %>% kable

```

```{r hrt_ba_winbugs_results}
hrt_bawb_results <- "
node	 mean	 sd	 MC error	2.5%	median	97.5%	start	sample
OR	0.8954	0.3737	0.01416	0.3843	0.8282	1.845	10001	10000
beta	-0.1301	0.1952	0.008616	-0.5204	-0.1325	0.2612	10001	10000
d	-0.1862	0.3861	0.01477	-0.9563	-0.1885	0.6126	10001	10000
prob.OR1	0.294	0.4556	0.01666	0.0	0.0	1.0	10001	10000
tau	1.082	0.778	0.05142	0.03262	0.9523	2.985	10001	10000
tau.bias	0.5373	0.09916	0.001401	0.3867	0.5227	0.7721	10001	10000
" %>% read.table(text=., sep="\t", header=TRUE, check.names=FALSE)

hrt_bawb_results %>% kable
```

## Hierarchical model

This is the code from Box 11.3 (p. 242). It was not included in the supplementary materials, so I typed it in by hand. I changed the name of a variable originally called 'mean' to 'my_mean', in case that is a reserved word in JAGS. This code runs, but I do not have official results to compare it to. The data values have not been double-checked. I added a second chain with arbitrary initial values, just to see that they both give similar distributions.

```{r HRT_hierarchical}

model_code <- "
model {
  # Randomized controlled trials
  for (i in 1:R){
    rct.prec[i] <- 1/(rct.serd[i] * rct.serd[i])
    rct.rd[i] ~ dnorm(rct.psi[i], rct.prec[i])
    rct.psi[i] <- theta[1] + (rct.z[i] * sd.theta[1])
    rct.z[i] ~ dnorm(0, 1)
  }
  
  # Comparative cohort studies model
  for (i in 1:C){
    coh.prec[i] <- 1/(coh.serd[i] * coh.serd[i])
    coh.rd[i] ~ dnorm(coh.psi[i], coh.prec[i])
    coh.psi[i] <- theta[2] + (coh.z[i] * sd.theta[2])
    coh.z[i] ~ dnorm(0, 1)
  }
  
  # Before after studies model
  for (i in 1:B){
    ba.prec[i] <- 1/(ba.serd[1] * ba.serd[1])
    ba.rd[i] ~ dnorm(ba.psi[i], ba.prec[i])
    ba.psi[i] <- theta[3] + (ba.z[1] * sd.theta[3])
    ba.z[i] ~ dnorm(0, 1)
  }
  
  # Combining all 3 sources
  for (i in 1:T){
    theta[i] <- my_mean + (u[i] * sd.mean)  # !!! in case 'mean' is a reserved word
    u[i] ~ dnorm(0, 1)
    sd.theta[i]  ~ dnorm(0, 0.1)I(0,)
    var.theta[i] <- sd.theta[i] * sd.theta[i]   # !!! not used?
    prec.theta[i] <- 1/(sd.theta[i] * sd.theta[i])
  }
  
  my_mean ~ dnorm(0, 0.01)
  sd.mean ~ dnorm(0, 0.1)I(0,)
  var.mean <- sd.mean * sd.mean # !!! not used ?
  prec.mean <- 1/(sd.mean * sd.mean)
  
}" %>% textConnection

data <- list(
  R=9, C=7, B=10, T=3,
  rct.rd=c( -10.51552, -2.028398, 4.115085, 6.479482, 0.0078509, 0.0, 2.247191,
            -5.817028, -3.984064),
  rct.serd=c(4.762193, 2.871006, 7.142432, 5.032322, 0.8079891, 8.058098,
             3.1075, 44.53912, 5.587013),
  ba.rd=c(-4.036327, 2.304048, -0.6941801, -3.186446,-7.431126, -1.458522,
          -4.036984, -1.613824, -1.461775, -0.1177738),
  ba.serd=c(2.242277, 3.579612, 0.6056279, 0.9381518, 2.121014, 0.5100973,
            1.072718, 0.6358061, 0.507642, 0.1981163),
  coh.rd=c(-1.41, -2.19, -4.34, -2.84, -2.53, -0.23, -0.46),
  coh.serd=c(1.433, 4.71, 1.914, 1.052, 3.081, 0.232, 0.123)
)

initial_values <- list(
  list( my_mean=-0.001, sd.mean=0.001, sd.theta=rep(0.001, 3)),
  list( my_mean= 0.1, sd.mean=0.1, sd.theta=rep(0.01, 3))
)

parameters_to_save <- c("rct.rd", "coh.rd", "ba.rd", "my_mean")


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

summary(results)

plot(results, parameters=c("my_mean"))

```

# Exercises

## `breast_screening`

This is the code for exercise 11.5 (p. 248)

Breast Screening - 3-level Hierarchical Model for Generalised Evidence Synthesis

Following Prevost et al Stat Med 2000;19 & Spiegelhalter et al (2003) Example 8.5

The WinBUGS code gave this error when I tried to run it in JAGS:
```
Error in rjags::jags.model(file = model.file, data = data, inits = inits,  : 
  RUNTIME ERROR:
Compilation error on line 7.

```

Here is a copy of the WinBUGS version:
```
model {
  for (j in 1:Nstud) {
    logRR[j] <- log(RR[j])
    P[j]  <- 1/(SE[j]*SE[j])
    # Likelihood for within-type model
    logRR[j] ~ dnorm(delta[j],P[j])
    delta[j] ~ dnorm(phi[Type[j]],prec.stud[Type[j]])
  }

  # Likelihood for between-type model
  for (i in 1:Ntype) {
    phi[i] ~ dnorm(d,prec)
    prec.stud[i] <- 1/psi.sq[i]
    psi.sq[i] <- psi[i]*psi[i]
    psi[i] ~ dnorm(0,8)I(0,)
    
    # Prior for psi is derived on basis that individual study effect is unlikely to 
    # deviate from type effect by more than 4 times or less than 1/4.
    # Therefore, upper 95% limit for psi = log(16)/(2*1.96) = 0.71.
    # Half-Normal(0,0.35^2) prior distribution has this property.
  
  } 

  # Priors
       
  d ~ dnorm(0,0.1)
  
  # Prior for d is derived on the basis that the overall effect is unlikely to 
  # to exceed 500 in favour of either screening or control.
  # Therefore, prior SD = log(500)/1.96 = 3.17 & precision = 1/(3.17^2) = 0.1.
  
  prec <- 1/tau.sq
  tau.sq <- tau*tau
  tau ~ dnorm(0,30)I(0,)
  
  # Prior for tau is derived on the basis that an specific type effect is unlikely 
  # to deviate from the overall pooled effect (d) by more than 2 or less than 1/2.
  # Therefore, upper 95%  limit for tau = log(2)/(2*1.96) = 0.018.
  # Half-Normal(0,0.18^2) prior distribution has this property.
   

  # Transformation to Relative Risk scale 
  for (i in 1:Ntype) {
    RR.type[i] <- exp(phi[i])
  }
  RR.ov <- exp(d)
}
```

To run it in JAGS, I moved the offending lines to a data block:

```{r breast_screening, eval=TRUE}

# Model

model_code <- "

data {
  for (j in 1:Nstud) {
    logRR[j] <- log(RR[j])
  }
}

model {
  for (j in 1:Nstud) {
    # logRR[j] <- log(RR[j])  # !!! moved to data block
    P[j]  <- 1/(SE[j]*SE[j])
    # Likelihood for within-type model
    logRR[j] ~ dnorm(delta[j],P[j])
    delta[j] ~ dnorm(phi[Type[j]],prec.stud[Type[j]])
  }

  # Likelihood for between-type model
  for (i in 1:Ntype) {
    phi[i] ~ dnorm(d,prec)
    prec.stud[i] <- 1/psi.sq[i]
    psi.sq[i] <- psi[i]*psi[i]
    psi[i] ~ dnorm(0,8)I(0,)
    
    # Prior for psi is derived on basis that individual study effect is unlikely to 
    # deviate from type effect by more than 4 times or less than 1/4.
    # Therefore, upper 95% limit for psi = log(16)/(2*1.96) = 0.71.
    # Half-Normal(0,0.35^2) prior distribution has this property.
  
  } 

  # Priors
       
  d ~ dnorm(0,0.1)
  
  # Prior for d is derived on the basis that the overall effect is unlikely to 
  # to exceed 500 in favour of either screening or control.
  # Therefore, prior SD = log(500)/1.96 = 3.17 & precision = 1/(3.17^2) = 0.1.
  
  prec <- 1/tau.sq
  tau.sq <- tau*tau
  tau ~ dnorm(0,30)I(0,)
  
  # Prior for tau is derived on the basis that an specific type effect is unlikely 
  # to deviate from the overall pooled effect (d) by more than 2 or less than 1/2.
  # Therefore, upper 95%  limit for tau = log(2)/(2*1.96) = 0.018.
  # Half-Normal(0,0.18^2) prior distribution has this property.
   

  # Transformation to Relative Risk scale 
  for (i in 1:Ntype) {
    RR.type[i] <- exp(phi[i])
  }
  RR.ov <- exp(d)
}
" %>% textConnection


# Data

data <- list(
  RR=c(0.70,0.66,0.79,0.80,0.64,0.80,0.80,0.26,0.30,0.24), 
  SE=c(0.17,0.11,0.23,0.19,0.27,0.11,0.045,0.48,0.43,0.30),
  Type=c(1,1,1,1,1,2,2,2,2,2), # (type: 1=RCTs, 2=Obs)
  Nstud=10,
  Ntype=2
)

# Starting/initial values: same as random effects model without sd
initial_values <- list(
  list(
    delta=c(0,0,0,0,0,0,0,0,0,0),
    phi=c(0,0),
    d=0,
    tau=1,
    psi=c(1,1)
  )
)


parameters_to_save <- c("RR.ov", "RR.type", "d", "phi", "psi", "tau")


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```

```
Results from WinBUGS
	node	 mean	 sd	 MC error	2.5%	median	97.5%	start	sample
	RR.ov	0.6518	0.1066	0.002882	0.4429	0.6503	0.8739	20001	10000
	RR.type[1]	0.6847	0.06534	0.002545	0.5622	0.682	0.8209	20001	10000
	RR.type[2]	0.6142	0.1066	0.003044	0.4014	0.6182	0.8172	20001	10000
	d	-0.4416	0.1667	0.004506	-0.8145	-0.4303	-0.1348	20001	10000
	phi[1]	-0.3834	0.09619	0.003775	-0.5759	-0.3827	-0.1974	20001	10000
	phi[2]	-0.5033	0.1814	0.00515	-0.9128	-0.4809	-0.2019	20001	10000
	psi[1]	0.1047	0.09434	0.003594	0.002431	0.07846	0.3512	20001	10000
	psi[2]	0.4653	0.1672	0.003304	0.1885	0.4476	0.8519	20001	10000
	tau	0.14	0.1053	0.002884	0.005694	0.1203	0.3903	20001	10000

```

## To do

* copy the data from 'efm RD.xls' in to R code.
* add the code from the Solutions to Exercises