---
title: 'Chapter 7: Evidence synthesis in a decision modelling framework.'
author: "Bob"
date: "2025-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')      # coda.samples
library('dplyr')
library('tidyr')
library('jagsUI')
library('mcmcplots')  # caterplot
library('coda')
```


# `caesarean_section_example.odc`

The code below fixes a problem with the original initial values from WinBUGS not working in JAGS. The first set of initial parameter values results in this error:
```
Error in `rjags::jags.model()`:
! Error in node ce.ratio
Invalid parent values
```
unless you change the values of `p1` and `beta` to 1e-6 instead of 0.

```{r caesarean_section_example_odc}

# Model

model_code <- "model
{

  # META-ANALYSIS ADJUSTING FOR BASELINE RISK

  for (i in 1:Num) 
  {
  	rc[i] ~ dbin(pc[i], nc[i])
  	rt[i] ~ dbin(pt[i], nt[i])	
  	logit(pc[i]) <- mu[i]
  	logit(pt[i]) <- mu[i] + delta[i] 
  	delta[i] <- delta.star[i] + beta * (mu[i]-(-1.09))
  	mu[i] ~ dnorm(0.0,1.0E-3)
  	delta.star[i] ~ dnorm(d, prec)
  }	
  
  d ~ dnorm(0.0,1.0E-6)
  OR <- exp(d)	
  d.pred ~ dnorm(d,prec)
  beta ~ dnorm(0.0,1.0E-6)
  
  tau ~ dunif(0,10)
  tau.sq <- tau*tau
  prec <- 1/tau.sq

  # DEVIANCE CONTRIBUTION

  for (i in 1:Num) 
  {      
    rthat[i] <- pt[i] * nt[i] 
    dev[i] <- 2 * (rt[i] * (log(rt[i])-log(rthat[i]))  +  (nt[i]-rt[i]) * (log(nt[i]-rt[i]) - log(nt[i]-rthat[i])))
  }
	
  sumdev <- sum(dev[])
  
  # DATA FOR DECISION MODEL	
  
  # Probability of wound infection without antibiotics
  rc1 ~ dbin(p1,nc1)
  p1 ~ dbeta(1,1)
  
  # Probability of wound infection with antibiotics
  logit(p2) <- logit(p1) + d + beta*(logit(p1)-(-1.09))
  # Length of stay in hospital with wound infection
  loswd ~ dnorm(mnloswd,precwd)  
  precwd <- 1/pow(sdloswd/sqrt(numwd),2)		
  # Length of stay in hospital without wound infection	
  losnwd ~ dnorm(mnlosnwd,precnwd)  
  precnwd <- 1/pow(sdlosnwd/sqrt(numnwd),2)	
  
  # Antibiotics administration costs	
  drugadmin ~ dunif(4,10)
  
  # QALYs - wound infection
  QALYs.wd <- ((QALYwd/365)*loswd)+((Fullhealth/365)*(fllwupdays-loswd))
  # QALYs - no wound infection
  QALYs.nwd <- ((QALYnwd/365)*losnwd)+((Fullhealth/365)*(fllwupdays-losnwd))
  
  
  # DECISION MODEL	
  
  # Cost with prophylactic antibiotics									
  cst.trt <- (1-p2) * ((antibiotic+drugadmin) * 3 + (losnwd*inptnwd)) + p2 * ((antibiotic+drugadmin) * 3 + (loswd * inptwd))
  # Cost without prophylactic antibiotics
  cst.ctl <- (1-p1) * (losnwd*inptnwd) + p1 * (loswd*inptwd)		
  # Difference in cost		
  diff.cost <- cst.trt - cst.ctl
  
  # Number of wound infections avoided using prophylactic antibiotics
  diff.wd <- (nc1 * (p1 - p2))
  
  # QALYs - with prophylactic antibiotics	
  QALYs.trt <- (1-p2) * QALYs.nwd+p2 * QALYs.wd
  # QALYs - without prophylactic antibiotics	
  QALYs.ctl <- (1-p1) * QALYs.nwd+p1 * QALYs.wd
  # Difference in QALYs	
  diff.QALYs <- (QALYs.trt - QALYs.ctl)
  
  # Probability using prophylactic antibiotics costs less than not using prophylactic antibiotics
  Q <- step(cst.ctl - cst.trt)
  
  # Cost effectiveness ratio
  ce.ratio <- diff.cost/diff.wd	
  cu.ratio <- diff.cost/diff.QALYs			
  
  # INCREMENTAL NET BENEFIT

  for (k in 1:M) 
  {
  	Rc1[k] <- (k-1) * 2000
  	INB.QALYs[k] <- Rc1[k] * diff.QALYs - diff.cost		
  	ProbCU[k] <- step(INB.QALYs[k])
  }
}
" %>% textConnection

# Data

data <- list(
  Num=35, 
  rc1=41, 
  nc1=486, 
  inptwd=163.03, 
  inptnwd=107.26, 
  mnloswd=8.8, 
  sdloswd=3.5, 
  mnlosnwd=6.7,  
  sdlosnwd=7.1, 
  numwd=41, 
  numnwd=445, 
  antibiotic=5.67, 
  M=21, 
  QALYwd=0.68, 
  QALYnwd=0.88, 
  Fullhealth=1, 
  fllwupdays=20
)

data_df <- "rt	nt	rc	nc
4	133	28	136
2	68	1	56
5	55	8	55
3	100	0	33
4	183	3	44
0	46	4	55
0	119	1	39
2	60	6	30
1	46	1	49
0	50	2	50
1	43	1	47
0	80	3	80
2	196	14	190
2	124	2	58
2	74	8	78
4	167	5	140
0	48	2	49
0	102	1	99
0	85	3	43
6	136	16	130
0	74	2	74
2	61	2	61
3	146	9	132
1	64	7	65
0	16	1	16
2	31	6	37
1	50	2	49
2	26	0	27
3	133	12	66
6	244	17	117
1	52	2	61
2	43	4	50
2	48	3	45
3	40	1	40
1	50	4	50" %>% read.delim(text=., sep='\t')

for (colname in names(data_df)){
  data[[colname]] <- data_df[[colname]]
}

# Starting/initial values

initial_values <- list(
  list(
    tau=1,
    delta.star=c(0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0, 0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0),
    d=0,
    mu=c(0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0,0,0,0,0,0, 0,0,0,0,0,0,0,0,0,0,  0,0,0,0,0),
    p1=1e-6, # 1e-6 works, 0.0 does not
    beta=1e-6  # 1e-6 works, 0.0 does not
  ),
  
  list(
    tau=0.1,
    delta.star=c(1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1),
    d=1, 
    mu = c(1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1,1,1, 1,1,1,1,1), 
    p1=0.5, 
    beta=1)
)

parameters_to_save <- c("ProbCU", "INB.QALYs", "Rc1", "diff.QALYs","diff.cost","OR",
                        "beta","ce.ratio","cu.ratio","p1","p2",
                        "sumdev","tau.sq")

# meta-analysis decision model
mad_results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

mad_results_summary <- summary(mad_results) %>% as.data.frame()

mad_results_summary

```

```{r plot_results}
plot(mad_results, parameters=c("ProbCU[1]", "INB.QALYs[1]", "diff.QALYs","diff.cost","OR",
                        "ce.ratio","cu.ratio","p1","p2") )
```


### Correlation Scatterplot (CE plane): diff.QALYs vs diff.cost

Here we re-sample the model to monitor cost-effectiveness parameters.

For an example using coda.samples, see the course notes from [Bayesian Analysis in JAGS: EMD chapter 6](https://public.wsu.edu/~jesse.brunner/classes/bio572/Lab7_Bayesian.html).


```{r plot_coda_data}

ce_coda_sample <- coda.samples(mad_results$model, variable.names = c("diff.QALYs", "diff.cost"), 
    n.iter = 10000)

# one set of results per chain?

library(ggplot2)
# coda_sample[[1]] %>% as.matrix %>% as.data.frame %>% 
#   ggplot(aes(x=diff.QALYs, y=diff.cost)) + geom_point()

coda_sample_2_df <- function(my_coda_sample){
  seq_along(my_coda_sample) %>% lapply(function(chain){
    df <- my_coda_sample[[chain]] %>% as.matrix %>% as.data.frame
    df['chain'] <- chain
    df
  }) %>% 
    bind_rows %>% 
    mutate(chain=factor(chain))
}

ce_df <- ce_coda_sample %>% 
  coda_sample_2_df %>% 
  sample_frac(1L) # randomly scramble chain order

XLIM = c(0, 7e-4)     # use same limits as chapter 3
YLIM = c(-75, 50)
LAMBDAS = 10000*(1:3)
ce_df %>%
  filter(chain == 1) %>%  # to compare to Ch3, which had only one chain
  ggplot(aes(x=diff.QALYs, y=diff.cost, col=chain)) + 
  geom_point(size=0.5, alpha=0.1) +
  geom_abline(slope=0, intercept=0, linetype='solid') +
  geom_abline(slope=LAMBDAS, intercept=0, linetype='dotted') +
  theme(legend.position="none") +
  labs(x = "Incremental QALYs", y = "Incremental costs", title="Cost-effectiveness plane for Caesarian section example with meta-analysis") + 
  coord_cartesian(xlim=XLIM, ylim=YLIM) + 
  annotate("text", x = XLIM[2], y = LAMBDAS * XLIM[2], 
           size=3, hjust=0.8, vjust=-0.25,
           label = paste("lambda ==", LAMBDAS), parse=TRUE)

ggsave("ce_scatterplot_ch7.pdf")  # compare plot to Ch.3.

```

### Cost-effectiveness

```{r economic_results}
# inb_vars <- c("INB", "ProbCE")
inb_vars <- c("INB.QALYs", "ProbCU")

inb_df <- coda.samples(mad_results$model, variable.names = inb_vars, 
    n.iter = 10000) %>% 
  coda_sample_2_df %>%
  group_by(chain) %>%
  mutate(iteration=row_number()) %>%
  ungroup() %>%
  as.data.frame

inb_df_long <- inb_df %>% 
  pivot_longer(
    cols=1:42,
    names_to=c("metric", "k"),
    names_pattern="(INB.QALYs|ProbCU)\\[(\\d+)\\]",
    values_to="value"
  ) %>%
  mutate(k=as.integer(k), lambda=2000*(k-1))
```

### Cost-effectiveness acceptability curve for Caesarian section example.

Figure 3.5 [also 3.6(c)]. WTP = "Willingness To Pay"

```{r CE_acceptability_curve, message=FALSE}

inb_df_long %>%
  group_by(metric, lambda) %>%
  filter(metric=="ProbCU") %>%
  summarize(P_CU=mean(value)) %>% 
  ggplot(aes(x=lambda, y=P_CU, group=metric)) + geom_line() + 
    ylim(0,1) + 
    ylab("Probability cost effective (for utility)") + 
    xlab("Lambda, WTP per additional QALY") +
    ggtitle("Cost-effectiveness acceptability curve")
  

```

### Expected net benefit (INB) and 95% interval for a range of lambda values.

Figure 3.6(b): Expected incremental net benefit and 95% interval for a range of lambda.

```{r expected_INB}

 # quantile(probs = c(2.5, 50, 97.5)/100)
inb_df_long %>%
  group_by(metric, lambda) %>%
  filter(metric=="INB.QALYs") %>%
  summarize(
    mean_INB=mean(value),
    lo_end=quantile(value, probs=0.025),
    hi_end=quantile(value, probs=0.975)) %>% 
  ggplot(aes(x=lambda, group=metric)) + 
    geom_line(aes(y=mean_INB)) + 
    geom_line(aes(y=lo_end), linetype="dashed") + 
    geom_line(aes(y=hi_end), linetype="dashed")

```

<h1><font color="red"> Finish these </font></h1>
### ScatterPlot of ProbCU vs Rc1
### Summary stats: ProbCU

```{r ProbCU_vs_Rc1}

# This makes 21 ProbCU columns and 21 Rc1 columns
mad_coda_df <- coda.samples(
  mad_results$model, 
  variable.names = c("ProbCU", "Rc1"),
  n.iter = 10000) %>% 
  coda_sample_2_df

library(tidyr)

mad_probcu_df <- mad_coda_df[,1:21] %>% 
  mutate(row_num = row_number()) %>%
  pivot_longer(
    cols=names(mad_coda_df)[1:21], # ProbCU[1]:ProbCU[21],
    names_to = c("id"),
    names_pattern = "ProbCU\\[(.+)\\]",
    values_to = "ProbCU"
  )


mad_rc1_df <- mad_coda_df[,22:42] %>% 
  mutate(row_num = row_number()) %>%
  pivot_longer(
    cols=names(mad_coda_df)[22:42],
    names_to = c("id"),
    names_pattern = "Rc1\\[(.+)\\]",
    values_to = "Rc1"
  )

mad_probcu_rc1_df <- merge(mad_probcu_df, mad_rc1_df)

mad_probcu_rc1_df %>% 
  mutate(ProbCU=factor(ProbCU)) %>%
  ggplot(aes(x=`ProbCU`, y=`Rc1`)) + geom_violin()

```



```{r summary_plots}

ProbCU <- mad_results_summary[1:21,]
with(ProbCU, {
  plot(mean, type='b', ylim=c(0,1))
})

```

# Code for Exercises

## `Model_1_decision_model_odc.txt`

This is the code for Exercise 3.1, a deterministic decision tree comparing the use of prophylactic neuaminidase inhibitors (NIs) to standard care for influenza. Exercise 7.1a asks you to combine this with the meta-analysis in `meta-analysis.odc`.

```{r model_1_decision_model}

model_code <- "model
{ 

  # DECISION MODEL
  costtrt <- ((1-p2) *(cdoc+cdrug) +p2* (cdoc+(cdrug)+cflu))
  costctl <- ((1-p1)*(cdoc)+p1*(cdoc+cflu))
  diff.cost <- costtrt-costctl

}" %>% textConnection

#DATA

data <- list(
  cdrug=118,    # cost of neuraminidase treatment
  cdoc=19,      # cost of visit to GP
  cflu=100,     # cost of treating flu
  p1=0.05,      # probability of contracting flu under standard care
  p2=0.03       # probability of contracting flu under NI prophylaxis				   
)

#RESULTS

results <- jags(
  data = data,
  parameters.to.save = c("cdrug", "cdoc", "cflu", "p1", "p2",
                         "costtrt", "costctl", "diff.cost"),
  model.file = model_code,
  n.chains = 1,
  n.iter=1
)


results

```

## `meta_analysis_odc.txt`

This code is for the NI Meta-Analysis that goes with Exercise 7.1a.

```{r meta_analysis_odc}

meta_analysis_model_code <- "model { 

  # META-ANALYSIS OF EFFECTIVENESS - ODDS RATIO (OR)

  for( j in 1 : Num ) 
    {
      rc[j] ~ dbin(pc[j], nc[j])
      rt[j] ~ dbin(pt[j], nt[j])
      logit(pc[j]) <- mu[j]
      logit(pt[j]) <- mu[j]+delta[j]
      delta[j] ~ dnorm(d,prec)	
      mu[j] ~ dnorm(0.0,1.0E-6)		
    }
	
  # PRIOR DISTRIBUTIONS
  
  d ~ dnorm(0.0,1.0E-6)
  tau ~ dunif(0,10)
  tau.sq <- tau*tau
  prec <- 1/tau.sq
  	
  # ODDS RATIO
  
  or <- exp(d)

}" %>% textConnection

# DATA
data <- list(
  rt=c(3,3,11,7,3),  nt=c(268,252,553,414,144),   	# Trial data - treatment arm
  rc=c(19,6,34,40,9),  nc=c(268,251,554,423,144),   # Trial data - standard care arm
  Num = 5						                                # number of studies in the meta-analysis
)

# INITIAL VALUES: This is a list of lists, one per chain.
initial_values <- list(
  list(tau=1, d=0, mu=c(0,0,0,0,0), delta=c(0,0,0,0,0))
)

meta_analysis_results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = c("d", "tau", "or"),
            model.file = meta_analysis_model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

summary(meta_analysis_results)

plot(meta_analysis_results)

```


## `Model 3 - CEAC.xls`

Exercise 7.2c asks you to paste the values for ProbCE into the yellow shaded cells of this spreadsheet to plot a graph that includes the cost-effectiveness acceptability curve.

# TO DO: Plot the graph for exercise 7.2c in R.

# Solutions to Exercises

## Question 7.1: Decision model + Meta-Analysis solution

```{r exercise_7_1}
model_code <- "
model
{ 

  # META-ANALYSIS OF EFFECTIVENESS - ODDS RATIO (OR)
  
  for (j in 1:Num) 
  {
    rc[j] ~ dbin(pc[j], nc[j])
    rt[j] ~ dbin(pt[j], nt[j])
    logit(pc[j]) <- mu[j]
    logit(pt[j]) <- mu[j]+delta[j]
    delta[j] ~ dnorm(d,prec)	
    mu[j] ~ dnorm(0.0,1.0E-6)	
  }
	
  # PRIOR DISTRIBUTIONS
  
  d ~ dnorm(0.0,1.0E-6)
  tau ~ dunif(0,10)
  tau.sq <- tau*tau
  prec <- 1/tau.sq
  
  # ODDS RATIO
  
  or <- exp(d)
  
  
  #CALCULATION OF p2 FROM ODDS RATIO
  
  logit(p2) <- logit(p1)+d
  
  #DECISION MODEL
  
  costtrt <- ((1-p2) *(cdoc+cdrug) +p2* (cdoc+(cdrug)+cflu))
  costctl <- ((1-p1)*(cdoc)+p1*(cdoc+cflu))
  diff.cost <- costtrt-costctl

}" %>% textConnection

#DATA

data <- list(
  rt=c(3,3,11,7,3),  nt=c(268,252,553,414,144),   # Trial data - treatment arm
  rc=c(19,6,34,40,9),  nc=c(268,251,554,423,144), # Trial data - standard care arm
  Num = 5,  			# number of studies in the meta-analysis
  cdrug=118,   		# cost of neuraminidase treatment
  cdoc=19, 				# cost of visit to GP
  cflu=100, 			# cost of treating flu
  p1=0.05					# prob of contracting flu under standard care
)

#INITIAL VALUES
initial_values <- list(
  list(tau=1, d=0, mu=c(0,0,0,0,0), delta=c(0,0,0,0,0))
)

parameters_to_save <- c("costtrt", "delta", "diff.cost", "mu", "or", "p2", "tau.sq")

ex1_results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000)

ex1_summary <- summary(ex1_results) %>% as.data.frame
ex1_summary['node'] <- row.names(ex1_summary)

ex1_summary
```
```{r ex1_density_plots}
densityplot(ex1_results, parameters=c('or', 'diff.cost'))
```
Comparing to the "official" answers from Solutions to Exercises, all the values plot on top of one another (except deviance, which does not appear in the official results).
```{r exercise_7_1_results_comparison}
official_results <- "node	 mean	 sd	 M_error	2.5pct	median	97.5pct	start	sample
costtrt	138.3	1.277	0.01573	137.5	138.2	139.6	5001	15000
delta[1]	-1.642	0.4318	0.00834	-2.655	-1.588	-0.9232	5001	15000
delta[2]	-1.268	0.4882	0.008834	-2.139	-1.315	-0.1579	5001	15000
delta[3]	-1.327	0.3072	0.006392	-1.904	-1.336	-0.6997	5001	15000
delta[4]	-1.638	0.3415	0.007694	-2.404	-1.609	-1.047	5001	15000
delta[5]	-1.384	0.4494	0.00709	-2.284	-1.395	-0.4277	5001	15000
diff.cost	114.3	1.277	0.01573	113.5	114.2	115.6	5001	15000
mu[1]	-2.649	0.2367	0.002346	-3.135	-2.641	-2.201	5001	15000
mu[2]	-3.625	0.3676	0.003682	-4.409	-3.603	-2.964	5001	15000
mu[3]	-2.709	0.1697	0.002022	-3.055	-2.705	-2.391	5001	15000
mu[4]	-2.301	0.1639	0.001936	-2.633	-2.296	-1.993	5001	15000
mu[5]	-2.709	0.3232	0.003224	-3.378	-2.692	-2.118	5001	15000
or	0.2645	0.42	0.004911	0.1042	0.2357	0.5086	5001	15000
p2	0.01346	0.01277	1.573E-4	0.005455	0.01225	0.02607	5001	15000
tau.sq	0.6589	2.799	0.05125	3.649E-4	0.1442	4.002	5001	15000" %>% read.delim(text=., sep="\t")

names(official_results) <- names(official_results) %>% gsub("X([0-9.]+)pct", "\\1%", .)

compare_cols <- c('node', 'mean', 'sd')
bind_rows(
  ex1_summary[-16, compare_cols] %>% mutate("source"='new'), 
  official_results[compare_cols] %>% mutate("source"='old')
) %>% ggplot(aes(x=node, y=mean, col=source, shape=source)) + 
  geom_point(size=4) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_shape_manual(values=c(1, 4))
```

