---
title: 'Chapter 10: Markov models'
author: "Bob"
date: "2025-03-02"
output: html_document
---

<font color="#FF0000">
NOTE: Several examples in this set are not working yet (these are chunks with eval=FALSE).
</font>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load_libraries, warning=FALSE, message=FALSE}
library('rjags')
library('coda')
library('jagshelper')
library('jagsUI')
library('mcmcplots')  # caterplot
library('dplyr')

library('tidyr') # pivot_wider
library('knitr') # to format table as HTML

```

# 'Example 10.2 revisited'

This model is hand-typed from p203 of the book. The challenge here is that we can no longer pretend that JAGS/BUGS code for data is the same as R code. That is true for lists and matrixes, but not for higher-dimensional arrays. So the data needs to be munged into the row-major order that JAGS expects. 

```{r example_10_2_revisited}

model_code <- "
model{
  for (tmt in 1:2) {    # tmt: 1=Seretide, 2=Fluticasone
    for (i in 1:4){     # four non-absorbing states
      r[tmt, i, 1:5] ~ dmulti(pi[tmt, i, 1:5], n[tmt, i]) # multinomial data
      pi[tmt, i, 1:5] ~ ddirch(prior[tmt, i, 1:5])        # Dirichlet prior
    }
  }

}
" %>% textConnection

# We need to do some munging to create an R data structure with the
# correct dimensions. First we make an R version of the `r` 3d array,
# using the data as it is printed in the text:
rr=structure(
  .Data=c(                    # Seretide
    210,	 60,	0,	1,	 1,	  # Row 1 (STW)
     88,	641,	0,	4,	13,	  # Row 2 (UTW)
      0,		0,	0,	0,	 0,	  # Row 3 (Hex)
      1,		0,	0,	0,	 1,	  # Row 4 (Pex)
                              # Fluticasone
     66,   32,	0,	0,	 2,		# Row 1 (STW)
     42,  752,	0,	5,	20,	  # Row 2 (UTW)
      0,    0,  0,	0,	 0,	  # Row 3 (Hex)
      0,    4,  0,	1,	 0    # Row 4 (Pex)
    ), 
    .Dim=c(5,4,2)             # Matrix dimensions # 2,4,5
)

# The Seretide matrix is t(r[,,1]), and the Fluticasone matrix is t(r[,,2])

# Now we need to shift the values around to the arrangement JAGS expects. Start with a 3d array full of placeholder values:
jr <- structure(
  .Data=rep(-1, 2*4*5), 
  .Dim=c(2,4,5)             # Matrix dimensions # 2,4,5
)

# Now copy in the numbers from the R array:
for (tmt in 1:2){
  for (i in 1:4){
    for (j in 1:5){
      jr[tmt, i, j] <- rr[j, i, tmt]
    }
  }
}

data <- list(
  r=jr,
  n=structure(                  # Row totals
    .Data=c(272,746,0,2,
            100,819,0,5), 
    .Dim=c(2,4)
  ),
  prior=structure(              # Dirichlet priors all = 1
    .Data=rep(1, 2*4*5), 
		.Dim=c(2,4,5)
	)
)

pi_results <- jags(
  data = data,
  parameters.to.save = "pi",
  model.file = model_code,
  n.chains = 1,
  n.adapt = 100,
  n.iter = 50000,
  n.burnin = 20000
)

```

The following chunk munges the results matrix to make a table more or less comparable to Table 10.4 in the text:

<style>
  table {
    width: auto;
  }
  td {
    border: 2px solid #999;
    padding: 10px 5px;
  }
  thead {
    background-color: #999;
    border: 2px solid #999;
    color: white;
  }
  tbody {
    color: black;
  }
</style>

```{r smry, results='asis'}
pi_smry <- summary(pi_results)

keeper_cols <- c("mean", "2.5%", "97.5%")

states <- c('STW', 'UTW', 'Hex', 'Pex', 'TF')
treatments = c('Seretide', 'Fluticasone')

seretide_flags <- dimnames(pi_smry)[[1]] %>% grepl("pi[1", ., fixed=TRUE)
fluticasone_flags <- dimnames(pi_smry)[[1]] %>% grepl("pi[2", ., fixed=TRUE)
# fluticasone_flags is not the same as !seretide_flags because of 'deviance'
tt_long <- rbind(
  pi_smry[seretide_flags, keeper_cols],
  pi_smry[fluticasone_flags, keeper_cols]
) %>% 
  as.data.frame %>% 
  mutate(
    value=sprintf("%0.3f<br/>(%0.2f, %0.2f)", mean, `2.5%`, `97.5%`),
  ) %>% select(value)

pat_matches <- row.names(tt_long) %>% regexec("pi\\[(\\d),(\\d),(\\d)\\]", .)
idx_df <- regmatches(row.names(tt_long), pat_matches) %>%
  sapply(function(v) v[2:4] %>% as.numeric) %>% 
  t %>%
  data.frame %>%
  setNames(., nm = c('tmt', 'i', 'j')) %>%
  mutate(
    Treatment = treatments[tmt],
    From = states[i],
    To = states[j]
  ) %>% 
  select(Treatment, From, To)

ttl <- bind_cols(idx_df, tt_long)

ttl %>% 
  pivot_wider(id_cols=c('Treatment', 'From'), names_from=c("To")) %>%
  as.data.frame %>% 
  # xtable %>% print(type = "html")
  knitr::kable(., format='html', escape=FALSE)

```

These numbers correspond well to those in Table 10.4, p 204. I have not included rows showing the probabilities of transitioning from the terminal state 'TF', since that state can only transition to itself.

# `asthma_1`

Estimating Transition Probabilities

<font color="#FF0000">
<pre>
Error in rjags::jags.model(file = model.file, data = data, inits = inits,  : 
  Error in node (r[1,2,1]/rhat[1,2,1])
Invalid parent values
<p/re>
</font>

```{r asthma_1, eval=FALSE}


# Model

model_code <- "model{
	for (tmt in 1:2){										# Treatments tmt=1 (Seretide), tmt=2 (Fluticasone)
		for (i in 1:4){									  # There are 4 non-absorbing health states
			r[tmt, i, 1:5] ~ dmulti(pi[tmt,i,1:5],n[tmt,i])		# Multinomial data
			pi[tmt, i, 1:5] ~ ddirch(prior[tmt,i,1:5])		# Dirichlet prior for probs.

			for (j in 1:5){
				rhat[tmt,i,j] <- pi[tmt,i,j] * n[tmt,i]
				dev[tmt,i,j] <- 2*r[tmt,i,j] * log(r[tmt,i,j]/rhat[tmt,i,j])
			}
			resdev[tmt,i] <- sum(dev[tmt,i,1:5])
		}
	}

resdevtot<- sum(resdev[1:2,1:4])
}

" %>% textConnection


# Data: same as in random effects model above
data <- list(
  r=structure(
    .Data=c(
      210,	60,	0,	1,	1,	
      88,	641,	0,	4,	13,	
       0,		0,	0,	0,	0,	
       1,		0,	0,	0,	1,	
      
      66,		32,	0,	0,	2,	
      42,	  752,	0,	5,	20,	
      0,		0,	  0,	0,	0,	
      0,		4,	  0,	1,	0), 
      .Dim=c(2,4,5)),
  n=structure(
    .Data=c(272,746,0,2,	100,819,0,5), 
    .Dim=c(2,4)),
  prior=structure(
    .Data=c(1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	
						1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1), 
		.Dim=c(2,4,5))
)


# Starting/initial values: same as random effects model without sd
initial_values <- list(
  #Inits
  list(pi=structure(
    .Data=c(.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1,	
  					.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1), 
  	.Dim=c(2,4,5))
  ),
  
  #Alternative Inits
  list(pi=structure(
    .Data=c(.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,
  					.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2),
  	.Dim=c(2,4,5))
  )
  
)


parameters_to_save <- c("r", "pi", "rhat") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```


# `asthma_2`

Propagating Uncertainty into a Decision Model


```{r asthma_2}


# Model

model_code <- "
model{
#Data analysis
	for (tmt in 1:2){										#Treatments tmt=1 (Seretide), tmt=2 (Fluticasone)
		for (i in 1:4){									#There are 4 non-absorbing health states
			r[tmt,i,1:5] ~ dmulti(pi[tmt,i,1:5],n[tmt,i])		#Multinomial data
			pi[tmt,i,1:5] ~ ddirch(prior[tmt,i,1:5])		#Dirichlet prior for probs.
		}
	}
#Calculating summaries from a decision model
	for (tmt in 1:2){ 
		for (i in 1:5){ s[tmt,i,1]<- equals(i,1) }  	#Initialise starting state: 1 in STW, 0 in all other states
		for (i in 1:4){  
			for (t in 2:13){
				s[tmt,i,t]<- inprod(s[tmt,1:4,t-1], pi[tmt,1:4,i]) 	# Run the model for 12 cycles. 
																			# s[tmt,i,t] = no. in state i at time t under treatment tmt
				}
			E[tmt,i]<- sum(s[tmt,i,2:13])							#Sum up time spent in state i
			}
		E[tmt,5]<- 12 - sum(E[tmt,1:4])							#Time in TF = 12 minus time in other states.
	}

	for (i in 1:5){
		D[i]<- E[1,i] - E[2,i]										#Additional time in state i under Seretide rather than FT
		prob[i]<- step(D[i])										#Indicates whether Seretide gives longer time in state i
	}
		
}

" %>% textConnection

data <- list(
r=structure(.Data=c(210,60,0,1,1,	88,641,0,4,13,	0,0,0,0,0,	1,0,0,0,1,	
							66,32,0,0,2,	42,752,0,5,20,	0,0,0,0,0,	0,4,0,1,0), 
				.Dim=c(2,4,5)),
n=structure(.Data=c(272,746,0,2,	100,819,0,5), .Dim=c(2,4)),
prior=structure(.Data=c(1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	
							1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1,	1,1,1,1,1), 
				.Dim=c(2,4,5))
)


initial_values <- list(
  #Inits
  list(pi=structure(.Data=c(.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1,	
  								.6,.1,.1,.1,.1,	.1,.6,.1,.1,.1,	.1,.1,.6,.1,.1,	.1,.1,.1,.6,.1), 
  				.Dim=c(2,4,5))
  ),
  
  #Alternative Inits
  list(pi=structure(.Data=c(.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,
  								.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2,	.2,.2,.2,.2,.2),
  				.Dim=c(2,4,5))
  )

  
)

parameters_to_save <- c("r", "pi", "s", "E") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```
# `evsyn_markov_book_1`

Synthesis of Studies Reporting Aggregate and Event History Data

<font color="#FF0000">
<pre>
Error in rjags::jags.model(file = model.file, data = data, inits = inits,  : 
  RUNTIME ERROR:
Compilation error on line 56.
Attempt to redefine node n[1]
</pre>
</font>

```{r evsyn_markov_book_1, eval=FALSE}
model_code <- "
model {

#MODEL

# Random effects model for logs of study-specific  rates  lamda1 of leaving state 1
for (s in 1:15) { loglam1[s] ~ dnorm(L,prec) 	#  Random effects model for lambda1
						 log(lambda1[s]) <- loglam1[s]
						gamma12[s]<-lambda1[s]*(1-rho)	#Study specific transition rates 1 to 2
						gamma13[s]<-lambda1[s]*rho	     #Study specific transition rates 1 to 3
						}   

# PRIORS

gamma23 ~ dexp(.001)                              # vague prior for 'fixed' transition rate 2 to 3
L ~ dnorm(0,.00001)                                        # vague prior for population mean loglamda1
tau ~ dunif(0,1)                                              # vague prior for sd for population loglamda1
prec <- 1/(tau*tau)  
rho ~ dbeta(1,1)                                         # vague prior for proportion going to state 2 from 1


#LIKELIHOOD 

#2 studies reported aggregated total no. of transitions 
	for (s in 1:2){
		for (i in 1:2){
			r.agg[s,i,1:3]~dmulti(pi.agg[s,i,1:3],n.agg[s,i])
		}
	}
	
	for (obs in 1:6){
		pi.agg[obs,1,1]<- exp(-(gamma12[study[obs]]+gamma13[study[obs]])*T[obs])
		pi.agg[obs,1,2]<- gamma12[study[obs]]*(exp(-gamma23 * T[obs])  - exp(- (gamma12[study[obs]]+gamma13[study[obs]])*T[obs] ) ) /((gamma12[study[obs]]+gamma13[study[obs]])- gamma23)
		pi.agg[obs,1,3]<- 1 - pi.agg[obs,1,1] - pi.agg[obs,1,2]
	
		pi.agg[obs,2,1]<- 0
		pi.agg[obs,2,2]<- exp(-gamma23)
		pi.agg[obs,2,3]<- 1 - pi.agg[obs,2,2]
	}

# 3 studies (4 data points) reported numbers number starting in state 1 and entering state 3 #during time T. Study 4 consists of 2 groups of individuals with different follow-up times, T
for (i in 1 : 4) {
       r.d[i] ~ dbin(pi.agg[(i+2),1,3],n.d[i]) 
      }

# 8 studies reporting event history data informing individual transitions
for (i in 1:2) {r12[i] ~dpois(mu[i+5])
                      log(mu[i+5]) <- log(E12[i])  +  log(gamma12[i+5])  }
for (i in 1:2) {r13[i] ~dpois(mu[i+7])
                      log(mu[i+7]) <- log(E13[i])  +  log(gamma13[i+7])   }
for (i in 1:4) {r23[i] ~dpois(mu[i+9])
                      log(mu[i+9]) <- log(E23[i])  +  log(gamma23)     }

# 2 studies reporting event history data on first tranistion from state 1
for (i in 1:2) {
       n[i] <- m12[i] + m13[i]
       n[i] ~ dpois(nu[i])                                     #   likelihood for Poisson rate
       log(nu[i]) <- log(E[i]) +  loglam1[i+13]  
       m13[i] ~ dbin(rho,n[i])                            # likelihood for binomial proportion going to state 3
      }

#Three different ways to summarise random effects model for lambda1, and form transition #rates. 

#1. Use population mean, L
g12[1] <- (1-rho) * exp(L)
g13[1] <-   rho * exp(L)

#2. Use predictive distribution for loglam1
loglam1.pred~dnorm(L,prec)
log(lambda1.pred)<-loglam1.pred
g12[2] <- (1-rho) * lambda1.pred
g13[2] <- rho * lambda1.pred

#3. Calibrate to study 15
g12[3] <- (1-rho) * lambda1[15]
g13[3] <- rho * lambda1[15]

#Transition probabilities over observation period T=1, for each RE summary
for (j in 1:3){
	P11[j]<-exp(-(g12[j]+g13[j]))
	P12[j]<-g12[j]*(exp(-gamma23)-exp(-(g12[j]+g13[j])))/(g12[j]+g13[j]-gamma23)
	P13[j]<- (1-P11[j]-P12[j])
	P22[j]<- exp(-gamma23)
	P23[j]<- 1-P22[j]
}

}

" %>% textConnection

# DATA LIST
data <- list(                                                                                           

#STUDIES 1-2
#2 studies reporting aggregate data for all states, for given observation cycle, T
	r.agg=structure(.Data=c(87,	4,	9,	
						    					0,	3,	1,
						
												210,	8,	15,
												0,		5,		3), .Dim=c(2,2,3)), 
	n.agg=structure(.Data=c(100,	4,
												233,	8), .Dim=c(2,2)), 

# STUDIES 3-5
#3 studies reporting no. transitions to death for given observation cycle, T
       r.d=c(38,15,11,60),      n.d=c(181,177,103,335) , 

#Observation time & study number for 6 aggregate data observations
T=c(1,1, 2,1.5,2,5), study=c(1,2,3,4,4,5),

# STUDIES 6-13
#  8 studies with event history data on individual transitions
   r12=c(8,20),               E12=c(120,620),
   r13=c( 12,44),            E13=c(140,677),  
   r23=c(9,12,5,6),         E23=c(34,35,15,25),

#STUDIES 14-15
# 2 studies with event history data on first transition from state 1
     m12=c(18,30),     m13=c(40,75),         E=c(380,1169) #, trailing comma was OK in WinBUGS     

)                                          
     
# INITIAL VALUES    

initial_values <- list(
  list(tau=.2,   rho=.4,    gamma23=.02,    L=-3,
       loglam1=c(.02,.02,.02,.02,.02,   .02,.02,.02,.02, .02,   .02,.02,.02,.02,.02),loglam1.pred=.02),
  
  list(tau=.1,   rho=.5,    gamma23=.08,    L=-2,
       loglam1=c(.01,.03,.01,.03,.01,   .01,.03,.01,.01, .03,   .03,.01,.02,.01,.03),loglam1.pred=.03)

)

parameters_to_save <- c("P11", "P12", "P13", "P22", "P23") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```


# `evsyn_markov_book_2`

Including Studies with Treatment Comparisons

<font color="#FF0000">
<pre>
Error in rjags::jags.model(file = model.file, data = data, inits = inits,  : 
  RUNTIME ERROR:
Compilation error on line 70.
Attempt to redefine node n[1]
</pre>
</font>

```{r evsyn_markov_book_2, eval=FALSE}

model_code <- "
model {

#MODEL

# Random effects model for logs of study-specific  rates  lamda1 of leaving state 1
for (s in 1:15) { loglam1[s] ~ dnorm(L,prec) 	#  Random effects model for lambda1
						 log(lambda1[s]) <- loglam1[s]
						gamma12[s]<-lambda1[s]*(1-rho)	#Study specific transition rates 1 to 2
						gamma13[s]<-lambda1[s]*rho	     #Study specific transition rates 1 to 3
						}   

# PRIORS

gamma23 ~ dexp(.001)                              # vague prior for 'fixed' transition rate 2 to 3
L ~ dnorm(0,.00001)                                        # vague prior for population mean loglamda1
tau ~ dunif(0,1)                                              # vague prior for sd for population loglamda1
prec <- 1/(tau*tau)  
rho ~ dbeta(1,1)                                         # vague prior for proportion going to state 2 from 1

logh~dnorm(0,.00001)							#Normal prior on log-hazard ratio
log(h)<- logh

#LIKELIHOOD 

#2 studies reported aggregated total no. of transitions 
	for (s in 1:2){
		for (i in 1:2){
			r.agg[s,i,1:3]~dmulti(pi.agg[s,i,1:3],n.agg[s,i])
			r.aggt[s,i,1:3]~dmulti(pi.aggt[s,i,1:3],n.aggt[s,i])
		}
	}

	for (s in 1:2){
		pi.aggt[s,1,1]<- exp(-(gamma12[s]*h+gamma13[s])*T[s])
		pi.aggt[s,1,2]<- gamma12[s]*h*(exp(-gamma23 * T[s])  
		- exp(-(gamma12[s]*h+gamma13[s])*T[s] ) ) /(gamma12[s]*h+gamma13[s] - gamma23 )
		pi.aggt[s,1,3]<- 1 - pi.aggt[s,1,1] - pi.aggt[s,1,2]
	
		pi.aggt[s,2,1]<- 0
		pi.aggt[s,2,2]<- exp(-gamma23)
		pi.aggt[s,2,3]<- 1 - pi.aggt[s,2,2]
	}

	for (obs in 1:6){
		pi.agg[obs,1,1]<- exp(-(gamma12[study[obs]]+gamma13[study[obs]])*T[obs])
		pi.agg[obs,1,2]<- gamma12[study[obs]]*(exp(-gamma23 * T[obs])  - exp(- (gamma12[study[obs]]+gamma13[study[obs]])*T[obs] ) ) /((gamma12[study[obs]]+gamma13[study[obs]])- gamma23)
		pi.agg[obs,1,3]<- 1 - pi.agg[obs,1,1] - pi.agg[obs,1,2]
	
		pi.agg[obs,2,1]<- 0
		pi.agg[obs,2,2]<- exp(-gamma23)
		pi.agg[obs,2,3]<- 1 - pi.agg[obs,2,2]
	}

# 3 studies (4 data points) reported numbers number starting in state 1 and entering state 3 #during time T. Study 4 consists of 2 groups of individuals with different follow-up times, T
for (i in 1 : 4) {
       r.d[i] ~ dbin(pi.agg[(i+2),1,3],n.d[i]) 
      }

# 8 studies reporting event history data informing individual transitions
for (i in 1:2) {r12[i] ~dpois(mu[i+5])
                      log(mu[i+5]) <- log(E12[i])  +  log(gamma12[i+5])  }
for (i in 1:2) {r13[i] ~dpois(mu[i+7])
                      log(mu[i+7]) <- log(E13[i])  +  log(gamma13[i+7])   }
for (i in 1:4) {r23[i] ~dpois(mu[i+9])
                      log(mu[i+9]) <- log(E23[i])  +  log(gamma23)     }

# 2 studies reporting event history data on first tranistion from state 1
for (i in 1:2) {
       n[i] <- m12[i] + m13[i]
       n[i] ~ dpois(nu[i])                                     #   likelihood for Poisson rate
       log(nu[i]) <- log(E[i]) +  loglam1[i+13]  
       m13[i] ~ dbin(rho,n[i])                            # likelihood for binomial proportion going to state 3
      }

#Three different ways to summarise random effects model for lambda1, and form transition #rates. 

#1. Use population mean, L
g12[1] <- (1-rho) * exp(L)
g13[1] <-   rho * exp(L)

#2. Use predictive distribution for loglam1
loglam1.pred~dnorm(L,prec)
log(lambda1.pred)<-loglam1.pred
g12[2] <- (1-rho) * lambda1.pred
g13[2] <- rho * lambda1.pred

#3. Calibrate to study 15
g12[3] <- (1-rho) * lambda1[15]
g13[3] <- rho * lambda1[15]

#Transition probabilities over observation period T=1, for each RE summary
for (j in 1:3){
	P11[j]<-exp(-(g12[j]+g13[j]))
	P12[j]<-g12[j]*(exp(-gamma23)-exp(-(g12[j]+g13[j])))/(g12[j]+g13[j]-gamma23)
	P13[j]<- (1-P11[j]-P12[j])
	P22[j]<- exp(-gamma23)
	P23[j]<- 1-P22[j]
}

}

" %>% textConnection

# DATA LIST
data <- list(                                                                                           

#STUDIES 1-2
#2 studies reporting aggregate data for all states, for given observation cycle, T
	r.agg=structure(.Data=c(87,	4,	9,	
						    					0,	3,	1,
						
												210,	8,	15,
												0,		5,		3), .Dim=c(2,2,3)), 
	n.agg=structure(.Data=c(100,	4,
												233,	8), .Dim=c(2,2)), 

#Studies 1-2 also report a treatment arm
	r.aggt=structure(.Data=c(140,	2,	14,	
						    					0,	2,	0,
						
												264,	5,	27,
												0,		3,		2), .Dim=c(2,2,3)), 
	n.aggt=structure(.Data=c(156,	2,
												296,	5), .Dim=c(2,2)), 

# STUDIES 3-5
#3 studies reporting no. transitions to death for given observation cycle, T
       r.d=c(38,15,11,60),      n.d=c(181,177,103,335) , 

#Observation time & study number for 6 aggregate data observations
T=c(1,1, 2,1.5,2,5), study=c(1,2,3,4,4,5),

# STUDIES 6-13
#  8 studies with event history data on individual transitions
   r12=c(8,20),               E12=c(120,620),
   r13=c( 12,44),            E13=c(140,677),  
   r23=c(9,12,5,6),         E23=c(34,35,15,25),

#STUDIES 14-15
# 2 studies with event history data on first transition from state 1
     m12=c(18,30),     m13=c(40,75),         E=c(380,1169)     

)

# INITIAL VALUES
initial_values <- list(
  list(tau=.2,   rho=.4,    gamma23=.02,    L=-3,
       loglam1=c(.02,.02,.02,.02,.02,   .02,.02,.02,.02, .02,   .02,.02,.02,.02,.02),loglam1.pred=.02,logh=0),
  
  list(tau=.1,   rho=.5,    gamma23=.08,    L=-2,
       loglam1=c(.01,.03,.01,.03,.01,   .01,.03,.01,.01, .03,   .03,.01,.02,.01,.03),loglam1.pred=.03,logh=-1)
)

parameters_to_save <- c("P11", "P12", "P13", "P22", "P23") # What else?


results <- jags(data = data,
            inits = initial_values,
            parameters.to.save = parameters_to_save,
            model.file = model_code,
            n.chains = length(initial_values),
            n.adapt = 100,
            n.iter = 50000,
            n.burnin = 20000,
            n.thin = 2)

summary(results)

```